---
title: "Women on the Ballot? Investigating Gendered Perceptions of the Emotional Viability of Women to Serve in Politics since 1974"
author: "Inessa De Angelis"
thanks: "Code and data are available at: https://github.com/InessaDeAngelis/Perceptions_of_Women_in_Politics "
date: "today"
date-format: "long"
abstract: "At current rates, the World Economic Forum estimates that it will take at least 100 years to reach gender parity in politics. Utilizing data from The General Social Survey (GSS), this paper analyzes responses between 1974 and 2022 to begin to understand how demographic factors, political views, and party identification impact perceptions of the emotional viability of women to serve in politics. This paper reveals that despite gendered media framing and other previously identified barriers, female politicians are largely perceived by people across a variety of political views and party identifications as having suitable emotional capabilities to hold elected office."
keywords:
  - "Gender stereotypes" 
  - "Gender roles"
  - "Women in politics"
  - "Surveys"
format: pdf
toc: true
number-sections: true
bibliography: references_publication.bib
csl: apsa.csl 
---

```{r}
#| include: false
#| warning: false
#| message: false

#### Workspace set-up ####
## Load packages ##
library(tidyverse)
library(knitr)
library(kableExtra)
library(here)
library(ggplot2)
library(formattable)
library(gtsummary)
library(marginaleffects)

## Read in cleaned data sets which do not include NAs ##
# Read in cleaned respondent info data #
cleaned_respondent_info <- read.csv(here::here("outputs/data/cleaned_respondent_info.csv"))
show_col_types = FALSE

# Read in cleaned women in politics data #
cleaned_women_in_politics <- read.csv(here::here("outputs/data/cleaned_women_in_politics.csv"))
show_col_types = FALSE

# Read in summarized voter statistics data #
summarized_political_preferences <- read.csv(here::here("outputs/data/summarized_political_preferences.csv"))
show_col_types = FALSE

## Read in cleaned data sets which include NAs ##
# Read in cleaned respondent info data #
cleaned_respondent_info1 <- read.csv(here::here("outputs/data/cleaned_respondent_info1.csv"))
show_col_types = FALSE

# Read in cleaned women in politics data #
cleaned_women_in_politics1 <- read.csv(here::here("outputs/data/cleaned_women_in_politics1.csv"))
show_col_types = FALSE

# Read in summarized voter statistics data #
summarized_political_preferences1 <- read.csv(here::here("outputs/data/summarized_political_preferences1.csv"))
show_col_types = FALSE

# Create merged data sets #
combined_data = 
  merge(
    cleaned_respondent_info,
    summarized_political_preferences,
    by = c("year", "id")
  )

finalized_data = 
  merge(
    combined_data,
    cleaned_women_in_politics,
    by = c("year", "id")
  ) |>
  mutate(
   year = as.character(year)
  )

# Create another merged data set with NAs included #
combined_data1 = 
  merge(
    cleaned_respondent_info1,
    summarized_political_preferences1,
    by = c("year", "id")
  )

finalized_data1 = 
  merge(
    combined_data1,
    cleaned_women_in_politics1,
    by = c("year", "id")
  ) |>
  mutate(
    year = as.character(year)
  )

```

# Introduction 
Despite changing norms and perceptions, women are largely still missing from elected office in the United States (U.S.) and across the world. In 2023, the U.S. elected a historic number of  women to the House of Representatives with 125 (or 29%) and 25 to the Senate (or 25%) [@rutgers]. The current Vice President, Kamala Harris, is the first woman to hold this office. Although the number of women elected to the highest levels of government in the U.S. is moving toward gender-parity, progress is slow. Women's credibility and perceived power as politicians is impacted by gendered framing by traditional media, priming members of the public to hold biased views about their qualifications [@deGeus]. Framing refers to how information is chosen, interpreted, and presented by the media, influencing the audience’s perception of it as important and shaping their own decision making procedures [@bashevkin]. Although it has not yet happened, previous research has shown that the public is open to voting for a female president and that gender stereotypes perpetuated by the media are playing less of a role in the construction of public image and voting behaviors [@StereotypingQs]. However, gender stereotypes continue to influence public perceptions of women in politics. 

To gain a further understanding of the impact of gender stereotypes and the perceived emotional viability of women to serve in politics, I examine responses by political views and party identification, using data obtained from the U.S. General Statistical Survey (GSS) from NORC [@gss] at the University of Chicago. These perceptions of women in politics are then analyzed and modeled using logistic regression to understand their assocations with demographic factors such as gender, as well as tracking the perceptions over time. It is hypothesized that respondents with more liberal political views and Democrat party identifications will disagree that men are emotionally better suited for politics than women and that respondents with more conservative political views and Republican party identifications will agree that men are emotionally better suited for politics than women. 

My analysis emphasizes that perceptions of women in politics have generally become more positive over time since the early 1970s, with a small dip in the late 1990s, 2006, and 2016. People who self-identify as Democrats or as having liberal political views in most cases disagree with the premise that men are emotionally better suited for politics then women, while people who self-identify as Republicans or as having conservative political views in some cases agree with the premise. The number of people with conservative views who agree that men are emotionally better suited for politics is not as high as anticipated, while people with Moderate political views and Other party identifications hold unexpected views about women in politics. My findings generally reinforce our understanding of the political divide on questions about the role of women in politics. 
	
In the remainder of this paper, I commence with the [Perceptions of Women in Politics] section which provides an overview of previous research and gaps, then in the [Data and methods] section I outline the nature of the data obtained, limitations, and cleaning procedures. In the [Results] section, I present my statistical models and trends found during the analysis process. Then, in the [Discussion] section, I provide further insights and future areas of study. Lastly, in the [Conclusion], I summarize the main findings.

# Perceptions of Women in Politics
Research analyzing public attitudes and biases toward women in politics in the U.S. and other democratic nations has many research precedents [@dolan; @StereotypingQs]. The traditional media, through mechanisms like framing and priming, focus on how issues are presented and the structural bias they uphold [@commsanddemocracy]. Specifically, “... the ways public problems are formulated for the media audience” continue to play a significant role in the portrayal of women in politics [@commsanddemocracy, p. 7]. Research published by @vanderpas2020 highlights how media framing of politics as a “masculine” sphere reinforces that women are not qualified or welcome, threatening the existence of current and future female politicians. @aaldering2020 suggest that voters and scholars alike need to cease evaluating female politicians on the basis of gendered, male politician stereotypes perpetuated by the media and instead offer evaluations on the basis of actual leadership skills. 

Public opinion polling analyzed by @dolanhansen reveals that while the public is aware of barriers facing women running for office, whether it be limited childcare or party fundraising structures, there is limited interest in working to remove these barriers. @dolanlynch go beyond the growing body of research into public perceptions and attitudes to understand voter behavior and decision making procedures when voters have the choice of women on the ballot. Their research reveals that there is a relationship, albeit not significant, between public attitudes towards women in politics and actually casting their ballot for female candidates [@dolanlynch]. 

Furthermore, previous research has focused on the gendered and partisan nature of women in politics, with  women traditionally having more liberal political views and Democrat party identifications, and with men having more conservative political views and Republican party identifications [@barnes]. However, moderate voters and people affiliated with "Other" political parties continue to be understudied in American politics, especially surrounding their views on issues such as the role of women in politics [@fowler; @davisbook]. Voters with moderate views, as illustrated by @davisbook, are conventionally understood as having moderate left-leaning or right-leaning views, not necessarily centrist, “middle of the road” views, as proposed by @moderates. 

The historical and contemporary availability of GSS data means it is often drawn on for measuring public attitudes toward various social, economic, and political issues, including evaluating perceptions of women in politics. Recently, scholars employ GSS data as a way to augment other national or specific research-creation survey data, often focusing on answers to demographic or economic questions. @welchsigelman used GSS data from the first three years the women in politics question was asked to determine which demographic factors were most predisposed to supporting women in politics, while @diekman reveal systematic bias and gender stereotypes lead to discrepancies when men were asked to reveal their more favorable positions on “female-stereotypical” policy positions. @jennings utilized GSS data to perform a longitudinal, cross-generational analysis of the “gender gap”, focusing on the dynamics of generational change and corresponding political views. However, no recent research has solely examined GSS data between the commencement of respondents being asked the women in the politics question in the 1970s and the most recent iteration of the survey in 2022. In particular, this paper fills existing gaps by solely employing GSS data to see how responses by gender, age, political views, and party identification vary over the span of nearly fifty years, while placing emphasis on the responses of Moderates and people affiliated with “Other” political parties. 

# Data and Methods
The data used in this paper was retrieved from the US General Social Survey (GSS) from NORC at the University of Chicago [@gss]. I retrieved both demographic data and data relating to the perceived emotional viability of women in politics, political party affiliation and identification, from 1974 to 2022.

For demographic data, I acquired responses for the variables `age` and `sex` which provide insight into respondents' age and gender. I then obtained data relating to the emotional viability of women in serve in politics, encompassed in the variables `fepol`, `fepolv`, and `fepolnv`. The survey question for all three variables asks *"Tell me if you agree or disagree with this statement: Most men are better suited emotionally for politics than are most women"* [@gssCodebook]. Going forward, I will refer to this as the "women in politics question".

I additionally retrieved data for the variables `polviews` (political views) and `partyid` (party identification). The political views survey question asks *"We hear a lot of talk these days about liberals and conservatives. I'm going to show you a seven-point scale on which the political views that people might hold are arranged from extremely liberal--point 1--to extremely conservative--point 7. Where would you place yourself on this scale?"* [@gssCodebook]. The party identification survey question asks *"Generally speaking, do you usually think of yourself as a Republican, Democrat, Independent, or what?"* [@gssCodebook].

## Survey Methodology and Data Limitations
Since 1972, GSS has administered in-person interviews to track public opinion. Due to the Covid-19 pandemic, data from the year 2020 was not recorded and instead spans the years of 2020 and 2021. This data was collected using new and updated methodologies, which changed which people were selected and how they completed the GSS questionnaire. 

The 2022 GSS bridges the methodology from both the traditional face-to-face data collection processes utilized between 1972 and 2018 and the web-based collection method introduced during the Covid-19 pandemic in 2021, using a mixed mode that includes face-to-face, web, and telephone surveys. Throughout 2022, the GSS conducted an experiment where the sample was divided into two parts. The first part encompassed people doing the face-to-face survey, while the second part included people doing the web and telephone survey. The content of the two parts of the data was not intended to be compared, instead comparing the two modes of collecting data. The sample divided in two for the purpose of comparing methodologies is visible in the `fepolv` and `fepolnv` variables selected for this paper. In 2021 and 2022, the GSS survey updated the variable name from `fepol` to `fepolv` and `fepolnv`, although the phrasing of the question remained the same. Despite these changes in methodology, GSS [@gssCodebook] intends for the 2022 survey to be comparable to the 2018 GSS. Although the majority of the data utilized for this study was obtained prior to the Covid-19 pandemic, the 2021 and 2022 responses may be impacted by the changes in methodology

Previous iterations of GSS selected respondents by quota (1972-1976), Kish grid with age ordering (1975-2018), and last birthday (2021). The 2022 GSS shifts to a Kish grid without age ordering, where adults in the household are selected by their order in the household, not age [@gssCodebook]. These changes to the selection of respondents, especially considering technological literacy among older populations may impact the survey results from 2021 and 2022. 

### "No Answer" and "Don't Know" Responses
When moving from an in-person interviewing process to web and mixed-method interviews, GSS shifted their approach to how to code uncertainty, indecision, or a refusal to answer from respondents. Prior to the Covid-19 pandemic, if a respondent opted not to answer a question, interviewers coded it as "No Answer" or "Don't Know". However, with the web-based survey in 2021, there was no interviewer to determine how to code a respondent's non-response. Instead of "No Answer" or "Don't Know", users could skip the question. "Skipped on web" indicates that users read a question but skipped it. "Skipped on web" responses are separated from “No Answer” which continues to be employed for the face-to-face and phone surveys [@gssCodebook].

The women in politics survey question continues to have a high "No Answer" rate, with 82% of survey respondents opting not to respond. @fig-NA_responses illustrates years in which the women in politics survey question was not asked (solely shown as "NA" or omitted) and years in which it was asked, separated into the three possible answers of "Agree", "Disagree", and "NA". Previous research suggests that social desirability bias impacts respondents' willingness to reveal their true beliefs about social issues such as the role of women in politics [@StereotypingQs], which is an identified limitation of this survey and is likely affecting the number of responses collected (further discussed in [Limitations]). 

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: fig-NA_responses
#| fig-cap: Distribution of "No Answer" responses to the Women in Politics Question by year

# Note to self: add this kind of breakdown for the other two Qs on pol view and party id.

# Code to manually change colours referenced from: https://stackoverflow.com/questions/36048033/manually-colouring-plots-with-scale-fill-manual-in-ggplot2-not-working

# Custom colours referenced from: https://www.datanovia.com/en/blog/awesome-list-of-657-r-color-names/

finalized_data1 |>
  ggplot(mapping = aes(x = year, fill = women_in_politics)) +
  geom_bar() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=90, size = 8)) +
  labs(x = "Year", y = "Number of Respondents", fill = "Women in politics") +
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(size = 6)) +
  theme(legend.title = element_text(size = 8)) +
  scale_fill_manual(values = c("red2", "dodgerblue3", "grey"))

```

## Data Cleaning and Preperations 
Data was collected, cleaned, and analyzed using the statistical programming software R [@r], using functions from `tidyverse` [@rTidyverse], `ggplot2` [@rGgplot2], `dplyr` [@rDplyr], `janitor` [@rJanitor], `KableExtra` [@rKableExtra], `knitr` [@rknitr], `labelled` [@rLabelled], `haven` [@rHaven], `Formattable` [@rFormattable], and `here` [@rHere]. 

After downloading and selecting the variables of interest from GSS [@gss], I performed data cleaning based on the variable definitions outlined in the GSS codebook [@gssCodebook]. I removed the years in which the questions asking whether men are emotionally better suited for politics than women, political views, and party identification were not asked (the raw data set would display answers for those years as NA). I updated the respondent demographic information data set to reflect the years omitted, given that these demographic questions are asked annually. 

The political views and party identification questions were asked slightly more frequently between 1974 and 1994, including in some years that the women in politics question was not. The responses from 1976, 1980, 1984, and 1987 were removed to ensure consistency between the two data sets. Similar to the women in politics question, the party identification and political views were asked bi-annually starting in 1994, except in 2020 due to Covid-19. 

### Respondent Demographic Information
During face-to-face interviews, interviewers traditionally coded the `sex` variable based on their observations. However, it was not possible to make this determination with telephone and web administered surveys, so respondents were asked their sex recorded at birth and their current gender identity. GSS then re-coded these responses into one variable: `sex`. 

During the data cleaning process, I renamed the `sex` column to `gender`, with value 1 updated to "Male", and value 2 updated to "Female" based on the GSS codebook [@gssCodebook]. There were no changes in how respondents were asked to provide their age.

### Perceptions of Women in Politics
Between 1974 and 2018, the variable for the women in politics question was `fepol`. However, with the implementation of the online only survey in 2021 and the mixed-methodology survey in 2022, `fepolv` and `fepolnv` were introduced to replace `fepol`. Consequently, employing the `rbind()` function which combines groups of rows together, I merged together the `fepol`, `fepolv`, and `fepolnv` responses into one new variable labelled `women_in_politics`. During the data cleaning process, I update value 1 to "Agree", and value 2 to "Disagree" based on the GSS codebook [@gssCodebook]

There appears to be no pattern or clear methodology behind why the women in politics question was asked some years over others, as sometimes the question was asked two or three years in a row, then not asked in a subsequent year, before repeating the sequence. Starting in 1994, the women in politics question was asked bi-annually, except in 2020 due to Covid-19 impacting the surveying process. 

### Political Preferences
During the data cleaning process, I renamed the `polviews` column to `political_views`, with the value 1 updated to "Extremely liberal", 2 updated to "Liberal", 3 updated to “Slightly liberal,” 4 updated to “Moderate”, 5 updated to “Slightly conservative”, 6 updated to “Conservative”, and 7 to “Extremely conservative” based on the GSS codebook [@gssCodebook].
 
Furthermore, during the data cleaning process I renamed the `partyid` column to `party_identification`. I then changed value 0 to “Strong Democrat", 1 to "Not Strong Democrat", 2 to "Independent, Close to Democrat", 3 to “Independent”, 4 to "Independent, Close to Republican", 5 to "Not Strong Republican", 6 to "Strong Republican", and 7 to “Other", based on the GSS codebook [@gssCodebook]. 

## Data Descriptions 
### Respondent Demographics
Between 1974 and 2022, 37,005 responses were recorded for GSS surveys which specifically asked the women in politics, political views, and party identification questions. Of these respondents, 20,609 (56%) identified as female, with 16,346 (44%) identifying as male, and 50 (0.1%) opting not to share their **gender**. Respondents largely disagreed that men are better emotionally suited then women to serve in politics. 74% of respondents, spanning 1974-2022, disagree with the question, while 26% agree. Women disagree with the question at a higher rate (42%), compared to men (32%), while women also agree with the question at a slightly higher rate than men (14% compared to 12%). 8 people (0.0%) agreed with the question without sharing their **gender**, while 42 people (0.1%) disagreed.  

@fig-women_in_pol_year_gender illustrates how perceptions of women in politics by **gender** have risen and fallen over time. We can see that both men and women agreed the most frequently with the premise that men are emotionally better suited for politics then women in 1975, 1977, and 1978 and disagreed the most frequently with this premise in 2021 and 2022. The 1985 and 1986 surveys witnessed an increase of both men and women who agree with the women in politics question, then another increase of women who agree with the question in the 1994, 1996, 1998, and 2000 surveys. The number of both men and women who agreed with the women in politics question rose again slightly in 2006 and 2016. Despite slight fluctuations in more recent years, the number of respondents who agree with the premise that men are emotionally better suited for politics has never reached the same levels as when GSS began asking this question in the 1970s. 

The number of respondents by **gender** who disagree with the women in politics question has fluctuated slightly less over time (see @fig-women_in_pol_year_gender). The 1994, 1996, 1998, and 2000 surveys saw an increase in the number of women and men who disagree with the premise that men are emotionally better suited for politics, along with 2006 and 2016. 

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: fig-women_in_pol_year_gender
#| fig-cap: Responses to the women in politics question by year and gender

finalized_data |>
  ggplot(mapping = aes(x = year, fill = gender)) +
  geom_bar() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=90, size = 6)) + 
  labs(
    x = "Year",
    y = "Number of respondents",
    fill = "Women in Politics"
  ) +
  facet_wrap(vars(women_in_politics)) +
  guides(x = guide_axis(angle = 90)) +
  theme(legend.position = "bottom") +
  scale_fill_brewer(palette = "Set1")

```

### Political Views
Responses to the women in politics question varies by **political views**, with people who identify as having more liberal views disagreeing, people with more conservative views agreeing, and people with more moderate views generally falling in the middle of liberal- and conservative-minded people (see @fig-polviews_all). Liberal views refers to anyone who self-identified as having **Slightly Liberal**, **Liberal**, or **Extremely Liberal** views. Conservative views accounts for anyone who self-identified as having **Slightly Conservative**, **Conservative**, or **Extremely Conservative** views. 

```{r}
#| message: false
#| echo: false
#| label: fig-polviews_all
#| fig-cap: Responses to the women in politics survey question over time by all political views

finalized_data |>
  mutate(political_views = fct_relevel(political_views, "Extremely Liberal", "Liberal", "Slightly Liberal", "Moderate",
          "Slightly Conservative", "Conservative", "Extremely Conservative")) |>
  ggplot(mapping = aes(x = year, fill = political_views)) +
  geom_bar() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=90, size = 8)) +
  labs(x = "Year", y = "Number of Respondents", fill = "Political Views") +
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(size = 6)) +
  theme(legend.title = element_text(size = 8)) +
  scale_fill_brewer(palette = "Set1")
```

1,965 people (18.6%) who self-identify as holding **liberal** views agree with the premise that men are emotionally better suited for politics, while 8,625 people (81.4%) disagree with the premise. Of the people who agree, 975 (9.2%) identify as female, 988 (9.3%) identify as male, and 2 (0.0%) opted not to identify their **gender**. Of those who disagree with the women in politics question, 4,999 people (47.2%) identify as female, 3,619 identify as male (34.2%), and 7 (0.1%) opted not to identify their **gender**.

4,015 people (32.2%) who self-identify as holding **conservative** views agree with the premise of the women in politics question, while 8,472 people (67.8%) who self-identify as holding conservative views disagree with the women in politics question. Of the people who agree, 2,069 (16.6%) identify as female, 1,943 (15.6%) identify as male, and 3 (0.0%) opted not to identify their **gender**. Of those who disagree with the women in politics question, 4,377 people (35.1%) identify as female, 4,079 identify as male (32.7%), and 16 (0.1%) opted not to identify their **gender**.

Lastly, among people who self-identify as holding **moderate** views, 3,523 people (25.3%) agree with the women in politics question while 10,405 people (74.7%) disagree. Of the people who agree, 2,074 (14.9%) identify as female, 1,446 (10.4%) identify as male, and 3 (0.0%) opted not to identify their **gender**. Of those who disagree with the women in politics question, 6,115 people (43.9%) identify as female, 4,271 identify as male (30.7%), and 19 (0.1%) opted not to identify their **gender**.

### Party Identification 
Responses to the question whether men are emotionally better suited for politics then women varies by political **party identification** (see @fig-partyid_all). People who self-identify as aligning with the Democrats disagree at a higher percent then people who self-identify as aligning with the Republicans. Aligning with the Democrats refers to anyone who self-identified in GSS as Strong Democrat, Not Strong Democrat, or Independent, Close to Democrat. Aligning with the Republicans accounts for anyone who self-identified as Strong Republican, Not Strong Republican, or Independent, Close to Republican. Independent refers to people who self-identify with neither the Democrats or Republicans and Other accounts for people who align with a third party or candidate. 

```{r}
#| message: false
#| echo: false
#| label: fig-partyid_all
#| fig-cap: Responses to the women in politics survey question over time by all political views

finalized_data |>
   mutate(party_identification = fct_relevel(party_identification, "Strong Democrat", "Not Strong Democrat", "Independent, Close to Democrat", "Independent", "Independent, Close to Republican", "Not Strong Republican", "Strong Republican", "Other")) |>
  ggplot(mapping = aes(x = year, fill = party_identification)) +
  geom_bar() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=90, size = 8)) +
  labs(x = "Year", y = "Number of Respondents", fill = "Party Identification") +
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(size = 6)) +
  theme(legend.title = element_text(size = 8)) +
  scale_fill_brewer(palette = "Set1")

```

4,146 people (23%) who self-identify as aligning with the **Democrats** agree with the premise that men are emotionally better suited for politics then women, while 13,869 people (77%) who self-identify as being affiliated with the Democrats disagree with the premise. Of the people who agree, 2,286 (12.7%) identify as female, 1,857 (10.3%) identify as male, and 3 (0.0%) opted not to identify their **gender**. Of those who disagree with the women in politics question, 8,278 people (46.0%) identify as female, 5,578 identify as male (31.0%), and 13 (0.1%) opted not to identify their **gender**.

4,121 people (31.8%) who self-identify as being affiliated with the **Republicans** agree with the women in politics question, while 8,853 people (68.2%) disagree with the question. Of the people who agree, 2,174 (16.8%) identify as female, 1,994 (15.0%) identify as male, and 3 (0.0%) opted not to identify their **gender**. Of those who disagree with the women in politics question, 4,586 people (35.3%) identify as female, 4,256 identify as male (32.8%), and 11 (0.1%) opted not to identify their **gender**.

1,128 people (20.8%) agree with the women in politics question while 4,294 people (79.2%) who self-identify as being **independent** from a political party disagree. The break down by gender of those who agree is 619 women (11.4%), 507 men (9.4%), and 2 (0.0%) opted not to share their **gender**. Of the people who disagree, 2,423 (44.7%) identify as female, 1,888 (34.2%) identify as male, and 16 (0.3%) opted not to share their **gender**.

Finally, 486 people (81.8%) who self-identify as being affiliated with a third party disagree with the women in politics question while 108 people (18.2%) agree with the question. Of the people who agree, 39 (6.6%) identify as female and 69 (11.6%) identify as male. The break down by gender of those who disagree is 204 women (34.3%), 280 men (47.1%), and 2 (0.3%) opted not to share their **gender**. 

# Results
This paper aims to evaluate the role of demographic factors, including gender, political views, and party identification on perceptions of whether women are emotionally suited for political office. I am especially interested in seeing whether the combination of the respondent's gender and political views and the respondent's gender and party identification impacts their overall perceptions of women in politics. 

Statistical modeling was executed in the statistical programming language R [@r], using `rstanarm` [@rstanarm], `gtsummary` [@gtsummary], `marginaleffects` [@marginal], and `modelsummary` [@modelsummary]. The model was fit in a Bayesian framework using `rstanarm` [@rstanarm]. I used the standard weakly informed prior distributions of 2.5 provided in the `rstanarm` package.

## Political Views
The first model considers whether a respondent agreed or disagreed with the premise that men are emotionally better suited for politics than women, as a function of their gender and political views.  

$$
\begin{aligned}
y_i|\pi_i & \sim \mbox{Bern}(\pi_i) \\
\mbox{logit}(\pi_i) & = \beta_0+\beta_1 \times \mbox{gender}_i + \beta_2 \times \mbox{political views}_i \\
\beta_0 & \sim \mbox{Normal}(0, 2.5)\\
\beta_1 & \sim \mbox{Normal}(0, 2.5)\\
\beta_2 & \sim \mbox{Normal}(0, 2.5)
\end{aligned}
$$ {#eq-gender_women_polviews}

Where $y_i$ is the respondent's views on women in politics and equal to 0 if they agree that men are emotionally better suited and 1 if they disagree, $\mbox{gender}_i$ is the gender of the respondent and $\mbox{political views}_i$ is the political views of the respondent.

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: Examining whether a respondent agrees with the premise that men are emotionally better suited for politics than women, based on their gender and political views
#| message: false
#| warning: false

gender_women_polviews <- readRDS(here::here("Outputs/model/gender_women_polviews.rds"))

tbl_regression(
  gender_women_polviews) |>
  as_gt()
```

The estimates displayed in @tbl-modelresults suggest that men are 8% less likely than women to disagree that men are emotionally better suited for politics than women, aligning with previous findings. When respondents who identified as having **Extremely Liberal** political views are compared to those with **Liberal** and **Slightly Liberal** views, **Liberals** are 32% less likely to support women in politics and **Slightly Liberals** are 53% less likely. When compared to respondents with **Extremely Liberal** political views, **Moderates** are 79% less likely to disagree that men are emotionally better suited for politics than women. People with **Slightly Conservative** views, when compared to people with **Extremely Liberal** views are 89% less likely to believe women are emotionally suited for politics, while people with **Conservative** and **Extremely Conservative** views are over 100% not likely to believe that women are emotionally suited for politics. These results align with previous research on views about women in politics by ideological divide, although the views of **Moderates** will be further examined in the [Discussion] section.

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresultsmarginal
#| tbl-cap: Examining whether a respondent agrees with the premise that men are emotionally better suited for politics than women, based on their gender and political views

slopes(gender_women_polviews, newdata = "median") |>
  select(term, contrast, estimate) |>
  kable(
    col.names = c("Term", "Contrast", "Estimate"),
    digits = 2, booktabs = TRUE
  )
```

@tbl-modelresultsmarginal investigates whether a respondent agrees with the women in politics question, based on their gender and political views. A larger difference was expected by respondent gender, given previous research, with the model estimating that when compared to women, men are just 2% less likely to disagree that men are emotionally better suited for politics [@barnes]. When people with **Extremely Liberal** political views are compared to respondents with **Liberal** views, **Liberals** are 4% less likely to disagree that men are emotionally better suited for politics. Respondents with **Slightly Liberal** views, when compared to those with **Extremely Liberal** views, are 7% less likely to disagree. The model emphasizes that the more Liberal a respondent's views are, the more likely they are to disagree with the premise that women are emotionally not suited for politics. Respondents who identify as having **Moderate** political views are an estimated 12% less likely than those with **Extremely Liberal** views to disagree that men are emotionally better suited for politics. Moreover, when compared to people with **Extremely Liberal** political views, respondents with **Slightly Conservative** views are 14% less likely to disagree. When people with **Conservative** and **Extremely Conservative** views are compared to respondents with **Extremely Liberal** views, they are 21% and 29% respectively less likely to disagree with the premise that women are not emotionally suited for politics compared to men. 

Although there are small quantitative differences between respondents with **Liberal** and **Slightly Liberal** views, when compared to respondents with  **Extremely Liberal** views, the differences between **Extremely Conservative**, **Conservative**, and **Slightly Conservative** views, when compared to respondents with **Extremely Liberal** views are more distinguishable. It is also interesting that people with **Moderate** views are more closely aligned with people with **Slightly Conservative** views over those with **Slightly Liberal** views, reinforcing the observations published by @davisbook. 

## Party Identification
The second model considers whether a respondent agreed or disagreed with the premise that men are emotionally better suited for politics than women, as a function of their gender and party identification. 

$$
\begin{aligned}
y_i|\pi_i & \sim \mbox{Bern}(\pi_i) \\
\mbox{logit}(\pi_i) & = \beta_0+\beta_1 \times \mbox{gender}_i + \beta_2 \times \mbox{party identification}_i \\
\beta_0 & \sim \mbox{Normal}(0, 2.5)\\
\beta_1 & \sim \mbox{Normal}(0, 2.5)\\
\beta_2 & \sim \mbox{Normal}(0, 2.5)
\end{aligned}
$$ {#eq-gender_women_partyid}

Where $y_i$ is the respondent's views on women in politics and equal to 0 if they agree that men are emotionally better suited and 1 if they disagree, $\mbox{gender}_i$ is the gender of the respondent and $\mbox{party identification}_i$ is the political views of the respondent.

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults_partyid
#| tbl-cap: Examining whether a respondent agrees with the premise that men are emotionally better suited for politics than women, based on their gender and party identification
#| message: false
#| warning: false

gender_women_partyid <- readRDS(here::here("Outputs/model/gender_women_partyid.rds"))

tbl_regression(
  gender_women_partyid) |>
  as_gt()
```

@tbl-modelresults_partyid investigates whether a respondent supports women in politics, on the basis of their gender and party identification. When compared to female respondents, male respondents are 10% less likely to disagree with the premise that men are emotionally better suited for politics than women. When respondents who identified as having **Strong Democrat** party identifications are compared to those with **Not Strong Democrat** and **Independent, Close to Democrat** views, **Not Strong Democrats** are 8% less likely to support women in politics and **Independent, Close to Democrats** are 31% more likely to support women in politics. **Independents** when compared to **Strong Democrats** are 17% more likely disagree that women are emotionally not suited for politics. Furthermore, when respondents who identified as having **Strong Democrat** party identifications are compared to those with **Independent, Close to Republican**, **Not Strong Republican**, and **Strong Republican** views, **Independent, Close to Republicans** are 13% less likely, **Not Strong Republicans** are 39% less likely, and **Strong Republicans** are 63% less likely to disagree that women are emotionally not suited for politics. People who identified their party affiliation as being **Other** are an estimated 35% more likely to support women in politics, than compared to people who are **Strong Democrats**. The predictions for Democrats and Republicans close to being Independent, Independents, and Others are particularly interesting as they reveal nuanced beliefs about the role of women in politics, unlike the other respondents who firmly fall along partisan lines. 

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresultsmarginal2
#| tbl-cap: Examining whether a respondent agrees with the premise that men are emotionally better suited for politics than women, based on their gender and party identification

slopes(gender_women_partyid, newdata = "median") |>
  select(term, contrast, estimate) |>
  kable(
    col.names = c("Term", "Contrast", "Estimate"),
    digits = 2, booktabs = TRUE
  )
```

@tbl-modelresultsmarginal2 highlights that despite significant partisan differences in views about women in politics, the quantitative differences may not be as stark as anticipated [@barnes]. When compared to female respondents, male respondents are 2% less likely to disagree with the premise that women are emotionally not suited for politics. **Strong Democrats** when compared to **Not Strong Democrats** are 1% less likely to agree with the women in politics questions, while those who identified as **Independent, Close to Democrat** are 5% more likely to disagree. Respondents with **Independent** views, when compared to **Strong Democrats** are an estimated 3% more likely to disagree with the women in politics question. Furthermore, all when compared to the views of **Strong Democrats**, **Independent, Close to Republicans** are 2% less likely, **Not Strong Republican** are 8%, and **Strong Republicans** are 13% less likely to disagree with the women in politics survey question. Lastly, respondents who identified themselves as **Other** are an estimated 8% more likely to disagree that women are emotionally not suited for politics, compared to **Strong Democrats**. The results in @tbl-modelresultsmarginal2 emphasize that despite partisan divides and views, perceptions of the emotional viability of women in serve in politics across the spectrum is better than anticipated. 

# Discussion
Despite gendered framing by the traditional media, the public largely sees women as being emotionally fit to serve as elected officials (74.3% of respondents believe women are emotionally fit). Support for women in politics by gender has risen and fallen over time, potentially responding to election candidates, socio-cultural phenomenons, and important news stories of the times (see @fig-women_in_pol_year_gender). As hypothesized, women disagree with the question of whether men are emotionally suited for politics than them at a higher rate, although the gap between genders is not as large as anticipated (see @tbl-modelresultsmarginal and @tbl-modelresultsmarginal2). With nearly three quarters of respondents disagreeing with the premise that men are more emotionally fit than women for politics, it is important to consider other factors that may be holding us back from achieving gender equality.  

## Perceptions of Women in Politics by Political Views 
The number of people with **Liberal** views who disagree with the question that men are emotionally better suited for politics then women is comparable to the hypothesized number, while the people with **Conservative** views who agree is not as high as anticipated, but still aligns with my hypothesis. My hypothesis did not account for people with **Moderate** views as the way people with moderate views feel about specific policy issues and survey questions is not consistent nor aligns with a single political party. However, in the case of the women in politics survey question, the people who identified as moderate aligned generally in the middle of people with liberal and conservative views, but leaned a little closer to respondents with **Slightly Conservative** views instead of **Slightly Liberal** views (see @tbl-modelresults and @tbl-modelresultsmarginal). The model's predictions align with @davisbook, who notes that moderate voters are not necessarily centrist but instead hold more moderate views on typically liberal and conservative policy perspectives. 

```{r}
#| message: false
#| eval: true
#| echo: false
#| label: fig-moderate_comp
#| fig-cap: Comparing Moderate and Liberal and Moderate and Conservative perceptions of women in politics over time
#| fig-subcap: ["Liberals & Moderates", "Conservatives & Moderates", "Including standard errors"]
#| layout-ncol: 1

# Code for sub-cap referenced from: https://github.com/RohanAlexander/telling_stories/blob/main/05-static_communication.qmd
# Panel (a)
finalized_data |>
  select(year, id, gender, political_views, women_in_politics) |>
  filter(political_views == "Moderate" | political_views == "Liberal" | political_views == "Slightly Liberal" | political_views == "Extremely Liberal") |>
   mutate(political_views = fct_relevel(political_views, "Extremely Liberal", "Liberal", "Slightly Liberal", "Moderate")) |>
  ggplot(mapping = aes(x = year, fill = political_views)) +
  geom_bar() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=90, size = 5)) + 
  labs(
    x = "Year",
    y = "Women in Politics",
    fill = "Political Views"
  ) +
  facet_wrap(vars(women_in_politics), scales = "free") +
  guides(x = guide_axis(angle = 90)) +
  theme(legend.text = element_text(size = 6)) +
  theme(legend.title = element_text(size = 9)) +
  theme(legend.position = "bottom") +
  scale_fill_brewer(palette = "Set1")

# Panel (b)
finalized_data |>
 select(year, id, gender, political_views, women_in_politics) |>
  filter(political_views == "Moderate" | political_views == "Conservative" | political_views == "Slightly Conservative" | political_views == "Extremely Conservative") |>
   mutate(political_views = fct_relevel(political_views, "Extremely Conservative", "Conservative", "Slightly Conservative", "Moderate")) |>
  ggplot(mapping = aes(x = year, fill = political_views)) +
  geom_bar() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=90, size = 5)) + 
  labs(
    x = "Year",
    y = "Women in Politics",
    fill = "Political Views"
  ) +
  facet_wrap(vars(women_in_politics), scales = "free") +
  guides(x = guide_axis(angle = 90)) +
  theme(legend.text = element_text(size = 6)) +
  theme(legend.title = element_text(size = 9)) +
  theme(legend.position = "bottom") +
  scale_fill_brewer(palette = "Set1")
```

### Perceptions of Women in Politics - Moderates versus Liberals and Conservatives 
Defining a "moderate" voter in American politics proves challenging and is a weakness of this survey (further discussed in [Limitations]). Scholars view moderates as not simply holding "middle of the road views", rather sharing moderate views of either liberals or conservatives on occasion if the social, economic, or political context aligns with their beliefs [@moderates; @davisbook]. Data from GSS reveals that people with moderate views fall in between people with liberal views and people with conservative views when both agreeing and disagreeing with the women in politics question (see @fig-moderate_comp). People with moderate views and people with liberal views hold somewhat similar views to the women in politics question (74.7% versus 81.4%). When broken down by gender, 47.2% of liberal-identifying women disagree with the women in politics question, while 43.9% of moderate-identifying women disagree. 34.2% of liberal-identifying men disagree, while 30.7% of moderate-identifying men disagree. Moreover, 9.2% of liberal-identifying women agree, while 14.9% of moderate-identifying women agree with the premise that men are emotionally better suited for politics. 9.3% of liberal-identifying men agree, while 10.4% of moderate-identifying men agree. In these circumstances, when analyzed by gender, the gap between people with liberal and moderate views closes slightly. 

In contrast, people with moderate views and people with conservative views also hold somewhat similar beliefs (74.7% versus 67.8%). When broken down by gender, 35.1% of conservative-identifying women disagree with the women in politics question, while 43.9% of moderate-identifying women disagree. 32.7% of conservative-identifying men disagree, while 30.7% of moderate-identifying men disagree. Moreover, 16.6% of conservative-identifying women agree, while 14.9% of moderate-identifying women agree with the premise that men are emotionally better suited for politics. 15.6% of conservative-identifying men agree, while 10.4% of moderate-identifying men agree. When analyzed by gender, the gap between people with conservative and moderate views aligns a little more closely, especially the number of women who agree with the premise that men are emotionally better suited for politics. A larger number of conservative-identifying men agree then moderate-identifying men with the women in politics question. 

While there is often significant disagreement between liberals, conservatives, and sometimes moderates in American politics, it is noteworthy that moderates generally hold "middle of the road views" on the women in politics question [@moderates].

```{r}
#| message: false
#| eval: true
#| echo: false
#| label: fig-other_comp
#| fig-cap: Comparing Democrat and Other party identifications and Republican  and Other party identifications perceptions of women in politics over time
#| fig-subcap: ["Democrats & Other", "Republicans & Other", "Including standard errors"]
#| layout-ncol: 1

# Code for sub-cap referenced from: https://github.com/RohanAlexander/telling_stories/blob/main/05-static_communication.qmd
# Panel (a)
finalized_data |>
  select(year, id, gender, party_identification, women_in_politics) |>
  filter(party_identification == "Strong Democrat" | party_identification == "Not Strong Democrat" | party_identification == "Independent, Close to Democrat" | party_identification == "Other") |>
   mutate(party_identification = fct_relevel(party_identification, "Strong Democrat", "Not Strong Democrat", "Independent, Close to Democrat", "Other")) |>
  ggplot(mapping = aes(x = year, fill = party_identification)) +
  geom_bar() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=90, size = 5)) + 
  labs(
    x = "Year",
    y = "Women in Politics",
    fill = "Party Identification"
  ) +
  facet_wrap(vars(women_in_politics), scales = "free") +
  guides(x = guide_axis(angle = 90)) +
  theme(legend.text = element_text(size = 6)) +
  theme(legend.title = element_text(size = 9)) +
  theme(legend.position = "bottom") +
  scale_fill_brewer(palette = "Set1")

# Panel (b)
finalized_data |>
 select(year, id, gender, party_identification, women_in_politics) |>
  filter(party_identification == "Strong Republican" | party_identification == "Not Strong Republican" | party_identification == "Independent, Close to Republican" | party_identification == "Other") |>
   mutate(party_identification = fct_relevel(party_identification, "Strong Republican", "Not Strong Republican", "Independent, Close to Republican", "Other")) |>
  ggplot(mapping = aes(x = year, fill = party_identification)) +
  geom_bar() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=90, size = 5)) + 
  labs(
    x = "Year",
    y = "Women in Politics",
    fill = "Political Views"
  ) +
  facet_wrap(vars(women_in_politics), scales = "free") +
  guides(x = guide_axis(angle = 90)) +
  theme(legend.text = element_text(size = 6)) +
  theme(legend.title = element_text(size = 9)) +
  theme(legend.position = "bottom") +
  scale_fill_brewer(palette = "Set1")

```

## Perceptions of Women in Politics by Party Identification 
Despite assumptions and previous research, the data illustrates that perceptions of the emotional viability of women to serve in politics matches political party stances (Democrat and Republican) on the matter, but the results are not as negative and that public is more supportive of women running for elected office then anticipated (see @tbl-modelresults_partyid and @fig-other_comp). The support of women in politics falls along party lines and is more positive than the ratios of women elected to the House of Representatives and the Senate from each party. Currently, there are 92 Democrat women (21%) and 33 Republican women (8%) elected to the House of Representatives (which has 435 members in total) [@rutgers]. There are 15 Democrat women (15%), 9 Republican women (9%), and 1 Independent woman (1%) elected to the Senate (which has 100 members in total) [@rutgers]. 

@tbl-modelresults_partyid shows that respondents who identify as **Independent, Close to Democrat** are an estimated 31% more likely to support women in politics, in comparison to **Strong Democrats**, which nearly matches the model's predictions for **Others** who are an estimated 35% more likely support women in politics. This is in contrast to **Independent, Close to Republicans** who are 13% less likely to support women in politics, when compared to **Strong Democrats**. The model's predictions are therefore suggesting that people with **Other** political views may be disillusioned Democrats or people who identify with third parties on the left of the political spectrum, instead of the right.  

The results from @tbl-modelresultsmarginal2, highlighting how **Others** are an estimated 8% more likely to disagree with the women in politics question than **Strong Democrats** leaves room for questions about the nature of respondents selecting this option. As highlighted by @fig-other_comp, respondents who identity with **Other** political parties and both agree and disagree with the women in politics question has risen in recent years, following a small decline in 2000, 2002, and 2004. Referencing previous research, one can assume that those who selected **Other** may have political views further to the left of the Democrats on the political spectrum, but without an opportunity for respondents to provide a short written response or interview, this option will remain ambiguous and a further area of study [@davisbook].  

## Limitations
Asking people about their political views and potential party affiliation remain sensitive topics in American society. As a result, some people may not be comfortable answering these questions which limits the diversity of potential responses collected by GSS. 

Starting from the assumption that respondents are likely to have political party views and party identifications that closely align with either of the two dominant political parties makes it challenging to collect accurate data about those who do not subscribe to either the Republican or Democrat playbooks [@davisbook]. Specifically, ambiguous survey question response options, such as **Moderate** (from the political views question) and **Other** (from the party identification question) should be expanded and clarified to ensure that the increasing number of people who find themselves aligning somewhere in the middle are not overlooked in the survey results [@fowler]. Understanding Moderate political views and Other party identifications is more nuanced than simply selecting "agree" or "disagree" on a survey, so without an opportunity for people to explain their views through an open-ended text box, important information which helps contextualize why respondents do or do not believe women are emotionally viable to serve in elected office is missed. 

Likewise, previous research has shown that people may have social desirability bias and mask their true beliefs when asked about topics such as the viability of women to serve in elected office [@StereotypingQs; @femquant]. Social desirability bias could be impacting the quality of the data collected, resulting in a significant number of "No Answer" responses, especially from respondents who complete the survey through the face-to-face and telephone modes. 

Despite the changing variable names to account for newer survey methods, the question asking whether men are emotionally better suited for politics than women hearkens back to the 1970s, regardless of the strides being made toward gender equality in politics. The phrasing of the question asking whether men are better emotionally suited is a leading question which is biased and primes the respondent to think about women in a specific way, reinforcing gender stereotypes [@femquant]. The media often frames women as being emotional, high strung, and not possessing “manly” leadership skills, whereas men are never asked about their emotions or have their behavior in the public eye covered through the lens of being emotional [@fallfromgrace; @bashevkin]. The same should be expected for the media coverage of women in politics and by extension, how the public is surveyed surrounding topics of women’s political  leadership and representation. GSS should consider asking respondents in future iterations of the survey about leadership qualities, relevant experience, and ability to address crucial policy issues. New phrasings of the women in politics survey question must account for the current progress being made toward gender equality, today's hybrid media environment, and the role that social media plays in disseminating political rhetoric and shaping the image of political candidates [@vanderpas2020]. 

Finally, GSS' changing sampling methodology over time may have lead to the imbalance of respondents by **gender**. This imbalance means that people who self-identify as female are represented at a slightly higher rate than people who self-identify as male, meaning readers should be aware when drawing conclusions about gendered perceptions of women in politics. Furthermore, other changes in methodology due to the Covid-19 pandemic may have impacted more recent responses, especially since access to reliable internet, cell phone service, and technological literacy are not a given among all respondents.

# Conclusion
Utilizing data from the U.S. General Social Survey, this paper analyzes and models perceptions of the emotional viability of women to serve in politics by political views and party identification over time from 1974 to 2022, in correlation with demographic factors such as gender and age. My results reveal that despite gendered media framing, the public is generally supportive of women seeking elected office. People self-identifying as Democrats and people with liberal and liberal-learning political views are more supportive of women in politics. However, the number of people who self-identify as Republicans and people with conservative and conservative-learning political views were more supportive of women in politics than anticipated. People who self-identify as having Moderate views generally hold "middle of the road" beliefs in response to the women in politics question, falling slightly closer to respondents with self-identified slightly conservative views. However, the views of Moderates and people affiliated with Other political parties, must be further studied in correlation with the women in politics question to develop a clearer understanding of the nuanced intersection of these topics.
 
Future studies should take into account the intersection of other demographic factors, political views, and party identifications while contextualizing responses to the women in politics question in the larger contemporary socio-cultural, economic, and political environment. In particular, future research should look at the intersection of race and identity, income, and highest level of education attained with these questions about political views, party identification, and the emotional viability of women to run for office. Future studies should also analyze political party structures, including candidate recruitment, the workplace culture in city halls, state legislatures, the House of Representatives, and the Senate, and the role of social media to help understand and correct course. Further emphasis should be placed on how methodologies, including survey questions are employed, ensuring gender-based stereotypes are not reinforced or the leadership roles of women in politics are not diminished [@femquant].

While media framing and corresponding public perceptions may not be as large of a problem as anticipated, we must work to understand all contributing factors to the low success rate of electing women to the top political offices in the U.S. and across the world. We cannot let gender equality take another 100 years of hard-fought victories to materialize [@WEF].

# References
