---
title: "Women on the Ballot? Investigating how the American Public Perceives the Emotional Viability of Women to Serve in Politics since 1974"
author: "Inessa De Angelis"
thanks: "Code and data are available at: https://github.com/InessaDeAngelis/Perceptions_of_Women_in_Politics "
date: "today"
date-format: "long"
abstract: "At current rates, the World Economic Forum estimates that it will take at least 100 years to reach gender parity in politics. Despite progress in recent years, why is it slated to take so long? Utilizing data from The General Social Survey (GSS), this paper analyzes responses from the American public between 1974 and 2022 to begin to understand how demographic factors, political views, and party identification impact perceptions of the emotional viability of women to serve in politics. This data reveals that despite gendered media framing, Americans largely perceive female politicians as having the emotional capabilities to hold elected office."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

#### Workspace set-up ####

## Load packages ##

library(tidyverse)
library(knitr)
library(kableExtra)
library(here)
library(ggplot2)
library(ggbeeswarm)
library(ggrepel)
library(formattable)

# Read in cleaned respondent info data #
cleaned_respondent_info <- read.csv(here::here("outputs/data/cleaned_respondent_info.csv"))
show_col_types = FALSE

# Read in cleaned women in politics data #
cleaned_women_in_politics <- read.csv(here::here("outputs/data/cleaned_women_in_politics.csv"))
show_col_types = FALSE

# Read in summarized voter statistics data #
summarized_political_preferences <- read.csv(here::here("outputs/data/summarized_political_preferences.csv"))
show_col_types = FALSE

# Create merged data sets #
combined_data = 
  merge(
    cleaned_respondent_info,
    summarized_political_preferences,
    by = c("year", "id")
  )

finalized_data = 
  merge(
    combined_data,
    cleaned_women_in_politics,
    by = c("year", "id")
  ) |>
  mutate(
   year = as.character(year)
  )

# Create filtered data sets for specific variables and responses #
liberal_data = 
  finalized_data |>
  select(year, id, gender, political_views, women_in_politics) |>
  filter(political_views == "Liberal" | political_views == "Slightly Liberal" | political_views == "Extremely Liberal")

conservative_data = 
  finalized_data |>
  select(year, id, gender, political_views, women_in_politics) |>
  filter(political_views == "Conservative" | political_views == "Slightly Conservative" | political_views == "Extremely Conservative")

moderate_data = 
  finalized_data |>
  select(year, id, gender, political_views, women_in_politics) |>
  filter(political_views == "Moderate")

moderate_liberal_data = 
  finalized_data |>
  select(year, id, gender, political_views, women_in_politics) |>
  filter(political_views == "Moderate" | political_views == "Liberal" | political_views == "Slightly Liberal" | political_views == "Extremely Liberal")

moderate_liberal_data = 
  finalized_data |>
  select(year, id, gender, political_views, women_in_politics) |>
  filter(political_views == "Moderate" | political_views == "Liberal" | political_views == "Slightly Liberal" | political_views == "Extremely Liberal") 

democrat_data =
  finalized_data |>
  select(year, id, gender, party_identification, women_in_politics) |>
  filter(party_identification == "Strong Democrat" | party_identification == "Not Strong Democrat" | party_identification == "Independent, Close to Democrat")

republican_data =
  finalized_data |>
  select(year, id, gender, party_identification, women_in_politics) |>
  filter(party_identification == "Strong Republican" | party_identification == "Not Strong Republican" | party_identification == "Independent, Close to Republican")

independent_data =
  finalized_data |>
  select(year, id, gender, party_identification, women_in_politics) |>
  filter(party_identification == "Independent")

other_data =
  finalized_data |>
  select(year, id, gender, party_identification, women_in_politics) |>
  filter(party_identification == "Other")

```

# Introduction 
Despite changing norms and perceptions, women are largely still missing from elected office in the United States (U.S.), Canada, and across the world. In 2023, the U.S. elected a historic number of  women to the House of Representatives with 125 (or 28.7%) and 25 to the Senate (or 25%) [@rutgers]. The current Vice President, Kamala Harris, is the first woman to hold this office. Although the number of women elected to the highest levels of government in the U.S. is inching toward gender-parity, progress is slow and hard fought. Women's credibility and perceived power as politicians is impacted by gendered framing by traditional media, priming members of the public to hold biased views about their qualifications [@deGeus]. Framing refers to how information is chosen, interpreted, and presented by the media, influencing the audience’s perception of it as important and shaping their own decision making procedures [@bashevkin]. Previous research has shown that the American public is open to voting for a female president and that gender stereotypes are playing less of a role in the construction of public image and voting behaviours [@StereotypingQs]. However, gender stereotypes continue to be a factor in perceptions of the viability of women to serve as elected officials. 

To gain a further understanding of the gendered stereotypes about the perceived emotional viability of women to serve in politics, I track responses by political views and party identification, using data obtained from the U.S. General Statistical Survey (GSS) from NORC [@gss] at the University of Chicago. These perceptions of women in politics are then analyzed to understand their correlations with demographic factors such as gender and age, as well as tracking the perceptions over time. 

My analysis emphasizes that perceptions of women in politics have generally improved over time since the early 1970s, with a small dip in the early 2010s. 
	
In the remainder of this paper, I commence with the [Data] section which outlines the nature of the data obtained, limitations, and cleaning procedures. In the [Results] section, I present trends found during the analysis process. Then, in the [Discussion] section, I provide further insights and future areas of study. In the [Conclusion], I summarize the main findings. Lastly, the [Appendix] contains a supplemental survey which proposes additional questions to study perceptions of women in politics. 

# Data
The data used in this paper was retrieved from the US General Social Survey (GSS) from NORC at the University of Chicago [@gss]. I obtained both demographic data and data relating to the perceived emotional viability of women in politics, political party affiliation and identification, from 1974 to 2022.

## Source Data

For demographic data, I obtained the following data, as outlined in @tbl-demographic.

| Variable  | New Name | Description      | Example |
|-----------|----------|------------------|---------|
| age       | age      | Respondent's age | 32      |
| sex       | gender   | Respondent's sex | female  |

: Demographic data obtained from GSS {#tbl-demographic}

As noted in @tbl-women_in_pol, I obtained this data relating to the emotional viability of women in serve in politics. The survey question for all three variables asks *"Tell me if you agree or disagree with this statement: Most men are better suited emotionally for politics than are most women"* [@gssCodebook]. Going forward, I will refer to this as the "women in politics question". A further discussion of these variables can be found later on in this section. 

| Variable  | New Name          | Description                                   | Example  |
|-----------|-------------------|-----------------------------------------------|----------|
| fepol     | women_in_politics | Men are better suited for politics than women | agree    |
| fepolv    | women_in_politics | Men are better suited for politics than women | disagree |
| fepolnv   | women_in_politics | Men are better suited for politics than women | disagree |

: Perceptions of Women in Politics data obtained from GSS {#tbl-women_in_pol}

Lastly, as outlined in @tbl-polviews, I retrieved data relating to political views and party identification. The political views survey question asks *"We hear a lot of talk these days about liberals and conservatives. I'm going to show you a seven-point scale on which the political views that people might hold are arranged from extremely liberal--point 1--to extremely conservative--point 7. Where would you place yourself on this scale?"* [@gssCodebook]. The party identification survey question asks *"Generally speaking, do you usually think of yourself as a Republican, Democrat, Independent, or what?"* [@gssCodebook].

| Variable | New Name             | Description                       | Example               |
|----------|----------------------|-----------------------------------|-----------------------|
| polviews | political_views      | Views ranked on a 7-point scale   | Slightly conservative |
| partyid  | party_identification | Republican, Democrat, Independent | Strong Democrat       |

: Political Party and Identification data obtained from GSS {#tbl-polviews}

## Data Limitations
Since 1972, GSS has been conducting in-person interviews to track public opinion. Due to the Covid-19 pandemic, the 2021 and 2022 surveys were conducted solely online and through mixed-methods. This change in methodology resulted in several limitations, which are discussed in further detail below.

### Survey Methodology 
The 2022 GSS bridges the methodology from both the traditional face-to-face data collection processes utilized between 1972 and 2018 and the web-based collection method introduced during the Covid-19 pandemic in 2021, using a mixed mode that includes face-to-face, web, and telephone surveys. Throughout 2022, the GSS conducted an experiment where the sample was divided into two parts. The first part encompassed people doing the face-to-face survey, while the second part included people doing the web and telephone survey. The content of the two parts of the data was not intended to be compared, instead comparing the two modes of collecting data. The sample divided in two for the purpose of comparing methodologies is visible in the `fepolv` and `fepolnv` variables selected for this paper. In 2021 and 2022, the GSS survey updated the variable name from `fepol` to `fepolv` and `fepolnv`, although the phrasing of the question remained the same. Despite these changes in methodology, GSS [@gssCodebook] intends for the 2022 survey to be comparable to the 2018 GSS. 

Previous iterations of GSS selected respondents by quota (1972-1976), Kish grid with age ordering (1975-2018), and last birthday (2021). The 2022 GSS shifts to a Kish grid without age ordering, where adults in the household are selected by their order in the household, not age [@gssCodebook]. These changes to the selection of respondents, especially considering technological literacy among older populations may impact the survey results from 2021 and 2022. 

### Gap in 2020 Data
Data from the year 2020 is not recorded and instead spans 2020 and 2021, which is inconsistent with the typically yearly GSS. There was also a shift in methodology, as previously discussed, to address risks from the Covid-19 pandemic which changed which people were selected and how they completed the GSS questionnaire. There was item non-response when respondents start the survey but some answers are missing (Stantcheva, p. 3).


### "No Answer" & "Don't Know" Responses
When moving from an in-person interviewing process to web and mixed-method interviews, GSS shifted their approach to how to code uncertainty, indecision, or a refusal to answer from respondents. Prior to the Covid-19 pandemic, if a respondent opted not to answer a question, interviewers coded it as "No Answer" and "Don't know". However, with the web-based survey in 2021, there was no interviewer to determine how to code a respondent's non-response. Instead of "No Answer" or "Don't know", users could skip the question. "Skipped on web" indicates that users read the questions but skipped it. "Skipped on web" responses are separated from “No Answer” which continues to be employed for the face-to-face and phone surveys [@gssCodebook].

### Impact on this study
Although the majority of the data utilized for this study was obtained prior to the Covid-19 pandemic, the 2021 and 2022 responses may be impacted by the changes in methodology and political context during and following the 2020 election. 2020 was a presidential election year in the United States, which saw Joe Biden and Kamala Harris (Democrat) beat Donald Trump and Mike Pence (Republican). Paired with a changing world and policy responses to the Covid-19 pandemic, whether it be mass-vaccination campaigns, stimulus cheques, or relief plans, respondents’ political views and party identification answers may be impacted. As well, Kamala Harris as a woman of colour on the Democrat ticket may have influenced responses to the women in politics survey question. Despite the significant methodological changes and political climate underpinning the 2021 and 2022 GSS, this will have less of an impact on long-term visualizations and trends.

## Data Cleaning
Data was collected, cleaned, and analyzed using the statistical programming software R [@r], using functions from `tidyverse` [@rTidyverse], `ggplot2` [@rGgplot2], `dplyr` [@rDplyr], `readr` [@rReadr], `tibble` [@rTibble], `janitor` [@rJanitor], `KableExtra` [@rKableExtra], `knitr` [@rknitr], `ggbeeswarm` [@rggbeeswarm], `ggrepel` [@rggrepel], `labelled` [@rLabelled], `haven` [@rHaven], `Formattable` [@rFormattable], and `here` [@rHere]. 

After downloading and selecting the variables of interest from GSS [@gss], I performed data cleaning based on the variable definitions outlined in the GSS codebook [@gssCodebook]. I removed the years in which the questions asking whether men are better suited emotionally for politics than women, political views, and party identification were not asked (the raw data set would display answers for those years as NA). I updated the respondent demographic information data set to reflect the years omitted, given that these demographic questions are asked every year. 

Between 1974 and 2018, the variable for the women in politics question was `fepol`. However, with the implementation of the online only survey in 2021 and the mixed-methodology survey in 2022, `fepolv` and `fepolnv` were introduced to replace `fepol`. Consequently, employing the `rbind()` function which combines groups of rows together, I merged together the `fepol`, `fepolv`, and `fepolnv` responses into one new variable labelled `women_in_politics`. There appears to be no pattern or clear methodology behind why the women in politics question was asked some years over others, as sometimes the question was asked two or three years in a row, then not asked in a subsequent year, before repeating the sequence. Starting in 1994, the women in politics question was asked bi-annually, except in 2020 due to Covid-19 impacting the surveying process. 

The political views and party identification questions were asked slightly more frequently between 1974 and 1994, including in some years that the women in politics question was not. The responses from 1976, 1980, 1984, and 1987 were removed to ensure consistency between the two data sets. Similar to the women in politics question, the party affiliation and political views were asked bi-annually starting in 1994, except in 2020 due to Covid-19. 

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-survey_statistics
#| tbl-cap: Summary statistics for the number of observations across the surveys from 1974 to 2022

#Code referenced from: https://github.com/christina-wei/Financial-Wellness-in-US/blob/main/outputs/paper/financial_satisfaction_in_us.qmd

finalized_data |>
  group_by(year) |>
  count() |>
  ungroup() |>
  summarise(
    mean = round(mean(n), digits = 0),
    max = max(n),
    min = min(n),
    sd = round(sd(n), digits = 0)
  ) |>
  kable(
    booktabs = TRUE,
    digits = 1
  )

```

## Respondent Demographic Information
During face-to-face interviews, interviewers traditionally coded the `sex` variable based on their observations. However, it was not possible to make this determination with telephone and web administered surveys, so respondents were asked their sex recorded at birth and their current gender identity. GSS then re-coded these responses into one variable: `sex`. 

During the data cleaning process, I renamed the `sex` column to `gender`, with the value 1 updated to "Male", and the value 2 updated to "Female" based on the GSS codebook [@gssCodebook].

Despite the changing survey methodology, respondents were still asked to identify their age in the web and mixed-method surveys. See @tbl-cleaned_respondent_info for a sample of the cleaned respondent demographic information.

```{r}
#| message: false
#| echo: false
#| label: tbl-cleaned_respondent_info
#| tbl-cap: Sample of Cleaned Respondent Demographic Information

head(cleaned_respondent_info) |>
  select(year, age, gender) |>
  kable(
    col.names = c("Year", "Age",	"Gender"),
  booktabs = TRUE
  )
  
```

## Perceptions of Women in Politics
Despite the changing variable names to account for newer survey methods, the question asking whether men are emotionally better suited for politics than women harkens back to the 1970s, despite the strides being made toward gender equality in American politics.  The phrasing of the question asking whether men are better emotionally suited is a leading question which is biased and primes the respondent to think about women in a specific way. Women are framed by traditional media and those on social media as being emotional, high strung, and not possessing “manly” leadership skills, so asking whether men are more emotionally suited over women reinforces traditional media framing and stereotypes [@fallfromgrace] [@bashevkin]. The media never asks men about their emotions or covers their behaviour in the public eye through the lens of being emotional and the same should be expected for the media coverage of women in politics.  GSS should consider asking future respondents in future iterations of the survey about leadership qualities, relevant experience, and ability to address crucial policy areas. Questions such as these are addressed in the supplement survey in the [Appendix]. 

Although this women in politics question is biased, the survey data can present interesting trends which will be discussed later in this paper. See @tbl-cleaned_women_in_politics for a sample of the cleaned data.  

```{r}
#| message: false
#| echo: false
#| label: tbl-cleaned_women_in_politics
#| tbl-cap: Sample of Cleaned Women in Politics Data

cleaned_women_in_politics |>
  slice(32000:32005) |>
  select(year, women_in_politics) |>
  kable(
    col.names = c("Year", "Women in Politics"),
  booktabs = TRUE
  )
  
```

## Political Preferences
During the data cleaning process, I renamed the `polviews` column to `political_views`, with the value 1 updated to "Extremely liberal", 2 updated to "Liberal", 3 updated to “Slightly liberal,” 4 updated to “Moderate”, 5 updated to “Slightly conservative”, 6 updated to “Conservative”, and 7 to “Extremely conservative” based on the GSS codebook [@gssCodebook].
 
Furthermore, during the data cleaning process I renamed the `partyid` column to `party_identification`. I then changed value 0 to “Strong Democrat", 1 to "Not Strong Democrat", 2 to "Independent, Close to Democrat", 3 to “Independent”, 4 to "Independent, Close to Republican", 5 to "Not Strong Republican", 6 to "Strong Republican", and 7 to “Other", based on the GSS codebook [@gssCodebook]. See @tbl-summarized_political_preferences for a sample of the cleaned data.  

```{r}
#| message: false
#| echo: false
#| label: tbl-summarized_political_preferences
#| tbl-cap: Sample of Summarized Political Preferences

summarized_political_preferences |>
  slice(1001:1006) |>
  select(year, political_views, party_identification) |>
  kable(
    col.names = c("Year", "Political Views", "Party Identification"),
  booktabs = TRUE,
  )
```

# Results
## Respondent Demographics
Between 1974 and 2022, there were 37,005 responses recorded for GSS surveys which specifically asked the women in politics, political views, and party identification questions. Of these respondents, 20,609 (55.7%) identified as female, with 16,346 (44.2%) identifying as male, and 50 (0.1%) opting not to share their **gender** (see @tbl-gender_summary).

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-gender_summary
#| tbl-cap: Number of respondents by gender 1974-2022

finalized_data |>
  group_by(gender) |>
  count() |>
  ungroup() |>
  mutate (pct = percent(n / sum(n), digit = 1)) |>
  kable(
    col.names = c("Gender", "Number of Responses", "Percentage"),
    booktabs = TRUE
  )
```

The age of survey respondents ranged from 18 to 89, with the highest number of responses per age falling within the 25-43 age range (see @tbl-age_summary). 

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-age_summary
#| tbl-cap: Age of Respondents

finalized_data |>
 summarize(
    youngest = min(age, na.rm = TRUE),
    oldest = max(age, na.rm = TRUE),
    average = mean(age, na.rm = TRUE)
  )
```

The breakdown of responses by age to the women in politics question is visible in @tbl-women_in_pol_age. 

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-women_in_pol_age
#| tbl-cap: Women in Politics by Age

finalized_data |>
  group_by(women_in_politics, age) |>
  count() |>
  ungroup() |>
  mutate (pct = percent(n / sum(n), digit = 1)) |>
  kable(
    col.names = c("Women in Politics", "Age", "Number of Responses", "Percentage"),
    booktabs = TRUE
  )
  
```

```{r}
#| message: false
#| echo: false
#| label: fig-women_in_pol_year_gender
#| fig-cap: Women in Politics by Year & Gender

#Code referenced from: https://tellingstorieswithdata.com/05-static_communication.html#bar-charts
finalized_data |>
  ggplot(mapping = aes(x = year, fill = gender)) +
  geom_bar() +
  theme_minimal() +
  labs(
    x = "Year",
    y = "Women in Politics",
    fill = "Gender"
  ) +
  facet_wrap(vars(women_in_politics), scales = "free") +
  guides(x = guide_axis(angle = 90)) +
  theme(legend.position = "bottom")

```

## Perceptions of Women in Politics 
Despite the biased wording of GSS' women in politics survey question, respondents largely disagreed that men are generally more emotionally suited then women to serve in politics. 74.3% of respondents, spanning 1974-2022, disagree with the question, while 25.7% agree. We can see in @tbl-women_in_pol_summary that women disagree with the question at a higher rate (41.9%), compared to men (32.3%). @tbl-women_in_pol_summary also highlights that women agree with the question at a slightly higher rate then men (13.8% compared to 11.8%). 8 people (0.0%) agreed with the question without sharing their **gender**, while 42 people (0.1%) disagreed. 

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-women_in_pol_summary
#| tbl-cap: Women in Politics

finalized_data |>
  group_by(women_in_politics, gender) |>
  count() |>
  ungroup() |>
  mutate (pct = percent(n / sum(n), digit = 1)) |>
  kable(
    col.names = c("Women in Politics", "Gender", "Number of Responses", "Percentage"),
    booktabs = TRUE
  )
```

## Political Views
Responses to the women in politics question varies by political views, with people who identify as having more liberal views disagreeing and people with more conservative views agreeing. Liberal views refers to anyone who self-identified in GSS as having `slightly liberal`, `liberal`, or `extremely liberal` views. Conservative views accounts for anyone who self-identified as having `slightly conservative`, `conservative`, or `extremely conservative` views. 

As outlined in @tbl-polviews_lib, we can see that 8,625 people (81.4%) who self identify as holding liberal views disagree with the women in politics question, while 1,965 people (18.6%) agree. @tbl-polviews_con highlights that 8,472 people (67.8%) who self identify as holding conservative views disagree with the women in politics question, while 4,015 people (32.2%) agree. 

Lastly, we can see in @tbl-polviews_moderate that among people who self identify as holding moderate views, 10,405 people (74.7%) disagree with the women in politics question, while 3,523 people (25.3%) agree. 

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-polviews_lib
#| tbl-cap: Responses to Women in Politics by Political Views- Liberal

liberal_data |>
  group_by(women_in_politics) |>
  count() |>
  ungroup() |>
  mutate (pct = percent(n / sum(n), digit = 1)) |>
  kable(
    col.names = c("Women in Politics", "Number of Responses", "Percentage"),
    booktabs = TRUE
  )

```

``` {r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-polviews_con
#| tbl-cap: Responses to Women in Politics by Political Views- Conservative

conservative_data |>
  group_by(women_in_politics) |>
  count() |>
  ungroup() |>
  mutate (pct = percent(n / sum(n), digit = 1)) |>
  kable(
    col.names = c("Women in Politics", "Number of Responses", "Percentage"),
    booktabs = TRUE
  )
```

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-polviews_moderate
#| tbl-cap: Responses to Women in Politics by Political Views- Moderate

moderate_data |>
  group_by(women_in_politics) |>
  count() |>
  ungroup() |>
  mutate (pct = percent(n / sum(n), digit = 1)) |>
  kable(
    col.names = c("Women in Politics", "Number of Responses", "Percentage"),
    booktabs = TRUE
  )
```

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-women_polviews_gender
#| tbl-cap: Responses to Women in Politics by Political Views & Gender

liberal_data |>
  group_by(women_in_politics, gender) |>
  count() |>
  ungroup() |>
  mutate (pct = percent(n / sum(n), digit = 1)) |>
  kable(
    col.names = c("Women in Politics", "Gender", "Number of Responses", "Percentage"),
    booktabs = TRUE
  )
```

## Party Identification 
Responses to the question whether men are more emotionally suited for politics then women varies by American political party affiliation. People who self-identify as aligning with the Democrats disagree at a higher percent then people who self-identify as aligning with the Republicans. Aligning with the Democrats refers to anyone who self-identified in GSS as `Strong Democrat`, `Not Strong Democrat`, or `Independent, Close to Democrat`. Aligning with the Republicans accounts for anyone who self-identified as `Strong Republican`, `Not Strong Republican`, or `Independent, Close to Republican`. `Independent` refers to people who self identify with neither the Democrats or Republicans and `Other` accounts for people who align with a third party or candidate. 

@tbl-partyid_dem reveals that 13,869 people (77%) who self identify as being affiliated with the Democrats disagree with the women in politics question while 4,146 people (23%) agree with the question. 

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-partyid_dem
#| tbl-cap: Responses to Women in Politics by Party Identification- Democrat

democrat_data |>
  group_by(women_in_politics) |>
  count() |>
  ungroup() |>
  mutate (pct = percent(n / sum(n), digit = 1)) |>
  kable(
    col.names = c("Women in Politics", "Number of Responses", "Percentage"),
    booktabs = TRUE
  )

```

As illustrated by @tbl-partyid_rep, 8,853 people (68.2%) who self identify as being affiliated with the Republicans disagree with the women in politics question while 4,121 people (31.8%) agree with the question. 

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-partyid_rep
#| tbl-cap: Responses to Women in Politics by Party Identification- Republican

republican_data |>
  group_by(women_in_politics) |>
  count() |>
  ungroup() |>
  mutate (pct = percent(n / sum(n), digit = 1)) |>
  kable(
    col.names = c("Women in Politics", "Number of Responses", "Percentage"),
    booktabs = TRUE
  )

```

@tbl-partyid_ind shows that 4,294 people (79.2%) who self identify as being Independent from a political party disagree with the women in politics question while 1,128 people (20.8%) agree with the question. 

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-partyid_ind
#| tbl-cap: Responses to Women in Politics by Party Identification- Independent

independent_data|>
  group_by(women_in_politics) |>
  count() |>
  ungroup() |>
  mutate (pct = percent(n / sum(n), digit = 1)) |>
  kable(
    col.names = c("Women in Politics", "Number of Responses", "Percentage"),
    booktabs = TRUE
  )

```

Finally, @tbl-partyid_other demonstrates that 486 people (81.8%) who self identify as being affiliated with a third party disagree with the women in politics question while 108 people (18.2%) agree with the question. 

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-partyid_other
#| tbl-cap: Responses to Women in Politics by Party Identification- Other

other_data |>
  group_by(women_in_politics) |>
  count() |>
  ungroup() |>
  mutate (pct = percent(n / sum(n), digit = 1)) |>
  kable(
    col.names = c("Women in Politics", "Number of Responses", "Percentage"),
    booktabs = TRUE
  )

```

# Discussion
Overall, the data reveals that despite gendered framing by traditional media that the American public largely sees women as being emotionally fit to serve as elected officials (74.3% of respondents in @tbl-women_in_pol_summary believe women are emotionally fit). With nearly three quarters of respondents disagreeing with the premise that men are more emotionally fit than women for politics, it is important to consider other factors that may be holding us back from gender equality.  

## Perceptions of Women in Politics by Demographic Factors  
As hypothesized, women disagreed with the question whether men are emotionally suited for politics then them at a high rate (see @tbl-women_in_pol_summary). It is noteworthy that women also agree with the question at a slightly higher rate then men (13.8% versus 11.8%)

## Perceptions of Women in Politics by Political Views 
The number of people with liberal views who disagree with the question that men are emotionally better suited for politics then women is comparable to the hypothesized number, while the people with conservative views who agree is not as high as anticipated, but still aligns with my hypothesis (see @tbl-polviews_moderate). My hypothesis did not account for people with moderate views as the way people with moderate views feel about specific policy issues and survey questions is not consistent nor align with a political party. However, in the case of the women in politics survey question, the people who identified as moderate aligned closely with the people with liberal views. 

### Perceptions of Women in Politics - Liberals vs Moderates 
Defining a "moderate" voter in American politics proves challenging and is a weakness of this survey (further discussed in [Limitations]). Scholars often view moderates as holding similar (but not identical) views to liberals on some issues. Data from GSS reveals in @fig-moderate_liberal that people with moderate views and people with liberal views hold very similar views to the women in politics question. While there is often significant disagreement between liberals, conservatives, and sometimes moderates in American politics, it is noteworthy that moderates align so closely with one faction on this question.

```{r}
#| message: false
#| echo: false
#| label: fig-moderate_liberal
#| fig-cap: Comparing Moderate & Liberal Perceptions of Women in Politics

moderate_liberal_data |>
  ggplot(mapping = aes(x = year, fill = political_views)) +
  geom_bar() +
  theme_minimal() +
  labs(
    x = "Year",
    y = "Women in Politics",
    fill = "Political Views"
  ) +
  facet_wrap(vars(women_in_politics), scales = "free") +
  guides(x = guide_axis(angle = 90)) +
  theme(legend.position = "bottom")
```

## Perceptions of Women in Politics by Party Identification 
Despite assumptions and previous research, this GSS data illustrates that perceptions of the emotional viability of women to serve in politics matches political party stances on the matter, but the results are not as negative and that public is more supportive of women running for elected office then anticipated. The support of women in politics falls along party lines and is more positive than the ratios of women elected to the House of Representatives and the Senate from each party. Currently, there are 92 Democrat women (21%) and 33 Republican women (8%) elected to the House of Representatives (which has 435 members in total) [@rutgers]. There are 15 Democrat women (15%), 9 Republican women (9%), and 1 Independent woman (1%) elected to the Senate (which has 100 members in total) [@rutgers]. 

## Limitations
Asking people about their political views and potential party affiliation remain sensitive topics, despite progress made since the 1970s. As a result, some people may not be comfortable answering these questions which limits the diversity in potential responses collected by GSS. 
Furthermore, previous research has shown that people may mask their true beliefs when asked about topics such as the viability of women to serve in elected office [@StereotypingQs], which could impact the quality of the data collected and number of "NA" and “don’t know” answers. Ambiguous survey question response options, such as `moderate` (from the political views question) should be clarified or eliminated to avoid respondents from making an uninformed decision, thus impacting the quality of the data. 

Politics and political views are much more nuanced then simply selecting "agree" or "disagree" on a survey, so without an opportunity for people to explain their views through an open-ended text box, important information which helps contextualize why they do or do not believe women are emotionally viable to serve in elected office is missed. 

The phrasing of the women in politics survey question, as previously discussed, does not account for current progress made toward gender equality, nor today's hybrid media environment and the role that social media plays in disseminating political rhetoric and shaping the image of political candidates. It is challenging to know the extent of which these factors impact GSS' data and should be accounted for in upcoming iterations of the survey. 

Finally, the change in methodology due to the Covid-19 pandemic may have impacted more recent responses, especially since access to reliable internet, cell phone service, and technological literacy are not a given among all people. 

## Future Research
### Supplemental Survey 
The acceptance of women as political leaders depends on several factors beyond their perceived emotional qualities. Therefore, I prepared a supplemental survey (available in the [Appendix]) to ask respondents about other factors that could impact their views and provide space for further contextualization. 

Previous research reveals that other perceived qualities about political candidates (especially female candidates) such as their leadership skills, highest level of education attained, and their ability to manage "serious" policy issues such as the economy can impact their electoral chances [@fallfromgrace]. Accordingly, the supplemental survey includes questions such as *"Do you think women possess the leadership skills required to serve in elected office?"*, *"Does the gender of a candidate matter more than their policies?"*, and *"What is the most important factoring when deciding which candidate to support in an election?"* I hypothesize that perceived leadership skills and the ability to manage important policy issues impact whether respondents decide to vote for female candidates. 

As well, I would like to find out more from respondents, through the question *"Is the current political climate supportive of women seeking elected office?"*, if the political climate and not the candidates are part of the problem with achieving gender equality. Responses to this question can also inform directions for future studies. 

### Possible Further Studies
There are many socio-economic, cultural, and demographic factors beyond an individual's age and gender that inform voting behaviours and perceptions of what constitutes a good political candidate. Consequently, future research should look at the intersection of race and identity, income, and highest level of education attained with these questions about political views, party identification, and the emotional viability of women to run for office.  

If media framing and political rhetoric are not solely shaping public attitudes towards women in politics, then what is causing the low success rate of electing women to office? Future studies should analyze political party structures, including candidate recruitment, the workplace culture in city halls, state legislatures, the House of Representatives, and the Senate, and the role of social media to help understand and correct course. 

A specific future study should investigate the intersection of political views, party identification, and the emotional viability of women to run for office questions with respondents' voting intentions and voting outcomes during the 2016 presidential election which saw Hillary Clinton (Democrat) face off against Donald Trump (Republican). The GSS variables `vote16`, `pres16`, and `if16who` could provide interesting data and insights [@gssCodebook]. 

# Conclusion
Utilizing data from the U.S. General Social Survey, this paper analyzed perceptions of the emotional viability of women to serve in politics by political views and party identification over time from 1974 to 2022, in correlation with demographic factors such as gender and age. My results reveal that despite gendered media framing that the American publicy is generally supportive of women seeking elected office. People identifying as Democrats and people with liberal and liberal-learning political views were more supportive of women in politics. However, the number of people who identifed as Republicans and people with conservative and conservative-learning political views were more supportive then anticipated. 

We cannot let gender equality take 100 years of hard-fought victories to materialize [@WEF].

# Appendix {.unnumbered}
The supplemental survey is available here: https://forms.gle/EFj72QiGS3fo186z5 

## Survey Preamble 
The General Social Survey collects information and maintains a historical record of respondents' attitudes, experiences, concerns, and practices. In order to strengthen the understanding of perceptions of women in politics, these questions are designed to better understand what socio-economic, cultural, and political factors and conditions contribute to a person's notion of whether women are fit to serve in elected political office, lead efforts to advance specific policy agendas, and increase the number of women elected to all levels of government.

This survey is voluntary and responses are anonymous. If you decide to participate, you can skip questions and withdraw at any time. Individual responses will be recorded in my data set using a user identification number.

Thank you for taking the time to complete this survey. Please contact inessa.deangelis@mail.utoronto.ca if you have questions or require any further information.

## Survey Questions
  (@) If you identify as a visible minority, what is your specific group?

  -Black
  
  -Chinese
  
  -Filipino
  
  -Japanese
  
  -Korean
  
  -South Asian-East Indian
  
  -Southeast Asian
  
  -North African 
  
  -Arab 
  
  -I do not identify as a visible minority
  
  -Other (please specify)
  

(@) What town or city do you live in?

(short answer text)


(@) What state do you live in?

(short answer text)


(@) How often do you engage with politics and political content?

  -Daily
  
  -A few times a week
  
  -A few times a month
  
  -Rarely 
  
  -Never
  
  
  (@) Where do you go to seek out political content?
  
  -Social media
  
  -Traditional media
  
  -Events
  
  -Other
  
  
  (@) Does the gender of a candidate matter more than their political affiliations?
  
  -Yes
  
  -No
  
  -Indifferent
  
  
  (@) Does the gender of a candidate matter more than their political ideology?
  
  -Yes
  
  -No
  
  -Indifferent
  
  
  (@) Does the gender of a candidate matter more than their policies?
  
  -Yes
  
  -No
  
  -Indifferent
  
 
  (@) What is the most important factoring when deciding which candidate to support in an election?
  
  -Gender
  
  -Personal characteristics
  
  -Leadership skills
  
  -Ideology
  
  -Political affiliation
  
  -Policies
  
  -Other
  
  
  (@) Do you think women possess the leadership skills required to serve in elected office?
  
  -Yes
  
  -No
  
  
(@) Do you think women can effectively manage political policies?
  
  -Yes
  
  -No
  

(@) Is the current political climate supportive of women seeking elected office?
  
  -Yes
  
  -No
  
  -Somewhat
  
## Submission Message
Thank you for taking the time to complete this survey. Your answers will help expand our understanding of perceptions of women in politics. 

# References
