---
title: "Women on the Ballot? Investigating how the American Public Perceives the Emotional Viability of Women to Serve in Politics since 1974"
author: "Inessa De Angelis"
thanks: "Code and data are available at: https://github.com/InessaDeAngelis/Perceptions_of_Women_in_Politics "
date: "today"
date-format: "long"
abstract: "At current rates, the World Economic Forum estimates that it will take at least 100 years to reach gender parity in politics. Despite progress in recent years, why is it slated to take so long? Utilizing data from The General Social Survey (GSS), this paper analyzes responses from the American public between 1974 and 2022 to begin to understand how demographic factors, political views, and party identification impact perceptions of the emotional viability of women to serve in politics. This data reveals XXX."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

#### Workspace set-up ####

## Load packages ##

library(tidyverse)
library(knitr)
library(kableExtra)
library(here)
library(ggplot2)
library(ggbeeswarm)
library(ggrepel)
library(formattable)

# Read in cleaned respondent info data #
cleaned_respondent_info <- read.csv(here::here("outputs/data/cleaned_respondent_info.csv"))
show_col_types = FALSE

# Read in cleaned women in politics data #
cleaned_women_in_politics <- read.csv(here::here("outputs/data/cleaned_women_in_politics.csv"))
show_col_types = FALSE

# Read in summarized voter statistics data #
summarized_political_preferences <- read.csv(here::here("outputs/data/summarized_political_preferences.csv"))
show_col_types = FALSE

combined_data = 
  merge(
    cleaned_respondent_info,
    summarized_political_preferences,
    by = c("year", "id")
  )

finalized_data = 
  merge(
    combined_data,
    cleaned_women_in_politics,
    by = c("year", "id")
  ) |>
  mutate(
   year = as.character(year)
  )

options(scipen = n)
```

# Introduction 
Despite changing norms and perceptions, women are largely still missing from elected office in the United States (U.S.), Canada, and across the world. In 2023, the U.S. elected a historic number of  women to the House of Representatives with 125 (or 28.7%) and 25 to the Senate (or 25%) [@rutgers]. The current Vice President, Kamala Harris, is the first woman to hold this office. Although the number of women elected to the highest levels of government in the U.S. is inching toward gender-parity, progress is slow and hard fought. Women's credibility and perceived power as politicians is impacted by gendered framing by traditional media, priming members of the public to hold biased views about their qualifications [@deGeus]. Framing refers to how information is chosen, interpreted, and presented by the media, influencing the audience’s perception of it as important and shaping their own decision making procedures [@bashevkin]. Previous research has shown that the American public is open to voting for a female president and that gender stereotypes are playing less of a role in the construction of public image and voting behaviours [@StereotypingQs]. However, gender stereotypes continue to be a factor in perceptions of the viability of women to serve as elected officials. 

To gain a further understanding of the gendered stereotypes about the perceived emotional viability of women to serve in politics, I track responses by political party affiliation and identification, using data obtained from the General Statistical Survey (GSS) from NORC [@gss] at the University of Chicago. These perceptions of women in politics are then analyzed to understand their correlations with demographic factors such as gender and age, as well as tracking the perceptions over time. 

My analysis emphasizes that perceptions of women in politics have generally improved over time since the early 1970s, with a small dip in the early 2010s. 
	
In the remainder of this paper, I commence with the Data section which outlines the nature of the data obtained, limitations, and cleaning procedures. In the Results section, I present trends found during the analysis process. Then, in the Discussion section, I provide further insights and future areas of study. In the Conclusion, I summarize the main findings. Lastly, the Appendix contains a supplemental survey which proposes additional questions to study perceptions of women in politics. 

# Data
The data used in this paper was retrieved from the US General Social Survey (GSS) from NORC at the University of Chicago [@gss]. I obtained both demographic data and data relating to the perceived emotional viability of women in politics, political party affiliation and identification, from 1974 to 2022.

## Source Data

For demographic data, I obtained the following data, as outlined in @tbl-demographic.

| Variable  | New Name | Description      | Example |
|-----------|----------|------------------|---------|
| age       | age      | Respondent's age | 32      |
| sex       | gender   | Respondent's sex | female  |

: Demographic data obtained from GSS {#tbl-demographic}

As noted in @tbl-women_in_pol, I obtained this data relating to the emotional viability of women in serve in politics. The survey question for all three variables asks *"Tell me if you agree or disagree with this statement: Most men are better suited emotionally for politics than are most women"* [@gssCodebook]. A further discussion of these variables can be found later on in this section. 

| Variable  | New Name          | Description                                   | Example  |
|-----------|-------------------|-----------------------------------------------|----------|
| fepol     | women_in_politics | Men are better suited for politics than women | agree    |
| fepolv    | women_in_politics | Men are better suited for politics than women | disagree |
| fepolnv   | women_in_politics | Men are better suited for politics than women | disagree |

: Perceptions of Women in Politics data obtained from GSS {#tbl-women_in_pol}

Lastly, as outlined in @tbl-polviews, I retrieved data relating to political views and party identification. The political views survey question asks *"We hear a lot of talk these days about liberals and conservatives. I'm going to show you a seven-point scale on which the political views that people might hold are arranged from extremely liberal--point 1--to extremely conservative--point 7. Where would you place yourself on this scale?"* [@gssCodebook]. The party identification survey question asks *"Generally speaking, do you usually think of yourself as a Republican, Democrat, Independent, or what?"* [@gssCodebook].

| Variable | New Name             | Description                       | Example               |
|----------|----------------------|-----------------------------------|-----------------------|
| polviews | political_views      | Views ranked on a 7-point scale   | Slightly conservative |
| partyid  | party_identification | Republican, Democrat, Independent | Strong Democrat       |

: Political Party and Identification data obtained from GSS {#tbl-polviews}

## Data Limitations
Since 1972, GSS has been conducting in-person interviews to track public opinion. Due to the Covid-19 pandemic, the 2021 and 2022 surveys were conducted solely online and through mixed-methods. This change in methodology resulted in several limitations, which are discussed in further detail below.

### Survey Methodology 
The 2022 GSS bridges the methodology from both the traditional face-to-face data collection processes utilized between 1972 and 2018 and the web-based collection method introduced during the Covid-19 pandemic in 2021, using a mixed mode that includes face-to-face, web, and telephone surveys. Throughout 2022, the GSS conducted an experiment where the sample was divided into two parts. The first part encompassed people doing the face-to-face survey, while the second part included people doing the web and telephone survey. The content of the two parts of the data was not intended to be compared, instead comparing the two modes of collecting data. The sample divided in two for the purpose of comparing methodologies is visible in the `fepolv` and `fepolnv` variables selected for this paper. In 2021 and 2022, the GSS survey updated the variable name from `fepol` to `fepolv` and `fepolnv`, although the phrasing of the question remained the same. Despite these changes in methodology, GSS [@gssCodebook] intends for the 2022 survey to be comparable to the 2018 GSS. 

Previous iterations of GSS selected respondents by quota (1972-1976), Kish grid with age ordering (1975-2018), and last birthday (2021). The 2022 GSS shifts to a Kish grid without age ordering, where adults in the household are selected by their order in the household, not age [@gssCodebook]. These changes to the selection of respondents, especially considering technological literacy among older populations may impact the survey results from 2021 and 2022. 

### Gap in 2020 Data
Data from the year 2020 is not recorded and instead spans 2020 and 2021, which is inconsistent with the typically yearly GSS. There was also a shift in methodology, as previously discussed, to address risks from the Covid-19 pandemic which changed which people were selected and how they completed the GSS questionnaire.

### Impact on this study
Although the majority of the data utilized for this study was obtained prior to the Covid-19 pandemic, the 2021 and 2022 responses may be impacted by the changes in methodology and political context during and following the 2020 election. 2020 was a presidential election year in the United States, which saw Joe Biden and Kamala Harris (Democrat) beat Donald Trump and Mike Pence (Republican). Paired with a changing world and policy responses to the Covid-19 pandemic, whether it be mass-vaccination campaigns, stimulus cheques, or relief plans, respondents’ political views and party identification answers may be impacted. As well, Kamala Harris as a woman of colour on the Democrat ticket may have influenced responses to the women in politics survey question. Despite the significant methodological changes and political climate underpinning the 2021 and 2022 GSS, this will have less of an impact on long-term visualizations and trends.

## Data Cleaning
Data was collected, cleaned, and analyzed using the statistical programming software R [@r], using functions from `tidyverse` [@rTidyverse], `ggplot2` [@rGgplot2], `dplyr` [@rDplyr], `readr` [@rReadr], `tibble` [@rTibble], `janitor` [@rJanitor], `KableExtra` [@rKableExtra], `knitr` [@rknitr], `ggbeeswarm` [@rggbeeswarm], `ggrepel` [@rggrepel], `labelled` [@rLabelled], `haven` [@rHaven], and `here` [@rHere]. 

After downloading and selecting the variables of interest from GSS [@gss], I performed data cleaning based on the variable definitions outlined in the GSS codebook [@gssCodebook]. I removed the years in which the questions asking whether men are better suited emotionally for politics than women, political views, and party identification were not asked (the raw data set would display answers for those years as NA). I updated the respondent demographic information data set to reflect the years omitted, given that these demographic questions are asked every year. 

Between 1974 and 2018, the variable for the women in politics question was `fepol`. However, with the implementation of the online only survey in 2021 and the mixed-methodology survey in 2022, `fepolv` and `fepolnv` were introduced to replace `fepol`. Consequently, employing the `rbind()` function which combines groups of rows together, I merged together the `fepol`, `fepolv`, and `fepolnv` responses into one new variable labelled `women_in_politics`. There appears to be no pattern or clear methodology behind why the women in politics question was asked some years over others, as sometimes the question was asked two or three years in a row, then not asked in a subsequent year, before repeating the sequence. Starting in 1994, the women in politics question was asked bi-annually, except in 2020 due to Covid-19 impacting the surveying process. 

The political views and party identification questions were asked slightly more frequently between 1974 and 1994, including in some years that the women in politics question was not. The responses from 1976, 1980, 1984, and 1987 were removed to ensure consistency between the two data sets. Similar to the women in politics question, the party affiliation and political views were asked bi-annually starting in 1994, except in 2020 due to Covid-19. 

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-survey_statistics
#| tbl-cap: Summary statistics for the number of observations across the surveys from 1974 to 2022

#Code referenced from: https://github.com/christina-wei/Financial-Wellness-in-US/blob/main/outputs/paper/financial_satisfaction_in_us.qmd

finalized_data |>
  group_by(year) |>
  count() |>
  ungroup() |>
  summarise(
    mean = round(mean(n), digits = 0),
    max = max(n),
    min = min(n),
    sd = round(sd(n), digits = 0)
  ) |>
  kable(
    booktabs = TRUE,
    digits = 1
  )

```

## Respondent Demographic Information
During face-to-face interviews, interviewers traditionally coded the `sex` variable based on their observations. However, it was not possible to make this determination with telephone and web administered surveys, so respondents were asked their sex recorded at birth and their current gender identity. GSS then re-coded these responses into one variable: `sex`. 

During the data cleaning process, I renamed the `sex` column to `gender`, with the value 1 updated to "Male", and the value 2 updated to "Female" based on the GSS codebook [@gssCodebook].

Despite the changing survey methodology, respondents were still asked to identify their age in the web and mixed-method surveys. The yougest survey respondents are 18 and the oldest are 89, with the average age of respondents being 47. See @tbl-cleaned_respondent_info for a sample of the cleaned respondent demographic information.

```{r}
#| message: false
#| echo: false
#| label: tbl-cleaned_respondent_info
#| tbl-cap: Sample of Cleaned Respondent Demographic Information

head(cleaned_respondent_info) |>
  select(year, age, gender) |>
  kable(
    col.names = c("Year", "Age",	"Gender"),
  booktabs = TRUE
  )
  
```

## Perceptions of Women in Politics
Despite the changing variable names to account for newer survey methods, the question asking whether men are emotionally better suited for politics than women harkens back to the 1970s, despite the strides being made toward gender equality in American politics.  The phrasing of the question asking whether men are better emotionally suited is a leading question which is biased and primes the respondent to think about women in a specific way. Women are framed by traditional media and those on social media as being emotional, high strung, and not possessing “manly” leadership skills, so asking whether men are more emotionally suited over women reinforces traditional media framing and stereotypes [@fallfromgrace] [@bashevkin]. The media never asks men about their emotions or covers their behaviour in the public eye through the lens of being emotional and the same should be expected for the media coverage of women in politics.  GSS should consider asking future respondents in future iterations of the survey about leadership qualities, relevant experience, and ability to address crucial policy areas. Questions such as these are addressed in the supplement survey in the Appendix. 

Although this women in politics question is biased, the survey data can present interesting trends which will be discussed later in this paper. See @tbl-cleaned_women_in_politics for a sample of the cleaned data.  

```{r}
#| message: false
#| echo: false
#| label: tbl-cleaned_women_in_politics
#| tbl-cap: Sample of Cleaned Women in Politics Data

cleaned_women_in_politics |>
  slice(32000:32005) |>
  select(year, women_in_politics) |>
  kable(
    col.names = c("Year", "Women in Politics"),
  booktabs = TRUE
  )
  
```

## Political Preferences
During the data cleaning process, I renamed the `polviews` column to `political_views`, with the value 1 updated to "Extremely liberal", 2 updated to "Liberal", 3 updated to “Slightly liberal,” 4 updated to “Moderate”, 5 updated to “Slightly conservative”, 6 updated to “Conservative”, and 7 to “Extremely conservative” based on the GSS codebook [@gssCodebook].
 
Furthermore, during the data cleaning process I renamed the `partyid` column to `party_identification`. I then changed value 0 to “Strong Democrat, 1 to "Not Strong Democrat", 2 to "Independent, Close to Democrat", 3 to “Independent”, 4 to "Independent, Close to Republican", 5 to "Not Strong Republican", 6 to "Strong Republican", and 7 to “Other", based on the GSS codebook [@gssCodebook]. See @tbl-summarized_political_preferences for a sample of the cleaned data.  

```{r}
#| message: false
#| echo: false
#| label: tbl-summarized_political_preferences
#| tbl-cap: Sample of Summarized Political Preferences

summarized_political_preferences |>
  slice(1001:1006) |>
  select(year, political_views, party_identification) |>
  kable(
    col.names = c("Year", "Political Views", "Party Identification"),
  booktabs = TRUE,
  )
```

# Results
## Respondent Demographics
Between 1974 and 2022, there were 37,0005 responses recorded for GSS surveys which specifically asked the women in politics, political views, and party identification questions. Of these respondents, 20,609 (55.7%) identifying as female, with 16,346 (44.2%) identifying as male, and 50 (0.1%) opting not to share their **gender** (see @tbl-gender_summary).

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-gender_summary
#| tbl-cap: Number of respondents by gender 1974-2022

finalized_data |>
  group_by(gender) |>
  count() |>
  ungroup() |>
  mutate (pct = percent(n / sum(n), digit = 1)) |>
  kable(
    col.names = c("Gender", "Number of Responses", "Percentage"),
    booktabs = TRUE
  )
```

The age of survey respondents ranged from 18 to 89, with the highest number of responses per age falling within the 25-43 age range (see @tbl-age_summary). 

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-age_summary
#| tbl-cap: Age of Respondents

finalized_data |>
  group_by(age) |>
  count() |>
  ungroup() |>
  mutate (pct = percent(n / sum(n), digit = 1)) |>
  kable(
    col.names = c("Age", "Number of Responses", "Percentage"),
    booktabs = TRUE
  )

finalized_data |>
 summarize(
    youngest = min(age, na.rm = TRUE),
    oldest = max(age, na.rm = TRUE),
    average = mean(age, na.rm = TRUE)
  )
```

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-polviews_summary
#| tbl-cap: Political Views

finalized_data |>
  group_by(political_views) |>
  count() |>
  ungroup() |>
  mutate (pct = percent(n / sum(n), digit = 1)) |>
  kable(
    col.names = c("Political Views", "Number of Responses", "Percentage"),
    booktabs = TRUE
  )

finalized_data |>
  group_by(party_identification) |>
  count() |>
  ungroup() |>
  mutate (pct = percent(n / sum(n), digit = 1)) |>
  kable(
    col.names = c("Party Identification", "Number of Responses", "Percentage"),
    booktabs = TRUE
  )

finalized_data |>
  group_by(women_in_politics) |>
  count() |>
  ungroup() |>
  mutate (pct = percent(n / sum(n), digit = 1)) |>
  kable(
    col.names = c("Women in Politics", "Number of Responses", "Percentage"),
    booktabs = TRUE
  )
```

# Discussion


# Conclusion


# Appendix {.unnumbered}
The supplemental survey is available here: https://forms.gle/EFj72QiGS3fo186z5 

## Survey Preamble 
The General Social Survey collects information and maintains a historical record of respondents' attitudes, experiences, concerns, and practices. In order to strengthen the understanding of perceptions of women in politics, these questions are designed to better understand what socio-economic, cultural, and political factors and conditions contribute to a person's notion of whether women are fit to serve in elected political office, lead efforts to advance specific policy agendas, and increase the number of women elected to all levels of government.

This survey is voluntary and responses are anonymous. If you decide to participate, you can skip questions and withdraw at any time. Individual responses will be recorded in my dataset using a user identification number.

Thank you for taking the time to complete this survey. Please contact inessa.deangelis@mail.utoronto.ca if you have questions or require any further information.

## Survey Questions

## Submission Message
Thank you for taking the time to complete this survey. Your answers will help expand our understanding of perceptions of women in politics. 

# References
