---
title: "Women on the Ballot? Investigating how the American Public Perceives the Emotional Viability of Women to Serve in Politics since 1974"
author: "Inessa De Angelis"
thanks: "Code and data are available at: https://github.com/InessaDeAngelis/Perceptions_of_Women_in_Politics "
date: "today"
date-format: "long"
abstract: "At current rates, the World Economic Forum estimates that it will take at least 100 years to reach gender parity in politics. Despite progress in recent years, why is it slated to take so long? Utilizing data from The General Social Survey (GSS), this paper analyzes responses from the American public between 1974 and 2022 to begin to understand how demographic factors, political views, and party identification impact perceptions of the emotional viability of women to serve in politics. This data reveals that despite gendered media framing, Americans largely perceive female politicians as having the emotional capabilities to hold elected office."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

#### Workspace set-up ####

## Load packages ##

library(tidyverse)
library(knitr)
library(kableExtra)
library(here)
library(ggplot2)
library(formattable)

# Read in cleaned respondent info data #
cleaned_respondent_info <- read.csv(here::here("outputs/data/cleaned_respondent_info.csv"))
show_col_types = FALSE

# Read in cleaned women in politics data #
cleaned_women_in_politics <- read.csv(here::here("outputs/data/cleaned_women_in_politics.csv"))
show_col_types = FALSE

# Read in summarized voter statistics data #
summarized_political_preferences <- read.csv(here::here("outputs/data/summarized_political_preferences.csv"))
show_col_types = FALSE

# Create merged data sets #
combined_data = 
  merge(
    cleaned_respondent_info,
    summarized_political_preferences,
    by = c("year", "id")
  )

finalized_data = 
  merge(
    combined_data,
    cleaned_women_in_politics,
    by = c("year", "id")
  ) |>
  mutate(
   year = as.character(year)
  )

# Create specific data sets for political views and party identifications #
liberal_data = 
  finalized_data |>
  select(year, id, gender, political_views, women_in_politics) |>
  filter(political_views == "Liberal" | political_views == "Slightly Liberal" | political_views == "Extremely Liberal")

conservative_data = 
  finalized_data |>
  select(year, id, gender, political_views, women_in_politics) |>
  filter(political_views == "Conservative" | political_views == "Slightly Conservative" | political_views == "Extremely Conservative")

moderate_data = 
  finalized_data |>
  select(year, id, gender, political_views, women_in_politics) |>
  filter(political_views == "Moderate")

moderate_liberal_data = 
  finalized_data |>
  select(year, id, gender, political_views, women_in_politics) |>
  filter(political_views == "Moderate" | political_views == "Liberal" | political_views == "Slightly Liberal" | political_views == "Extremely Liberal")

moderate_conservative_data = 
  finalized_data |>
  select(year, id, gender, political_views, women_in_politics) |>
  filter(political_views == "Moderate" | political_views == "Conservative" | political_views == "Slightly Conservative" | political_views == "Extremely Conservative") 

democrat_data =
  finalized_data |>
  select(year, id, gender, party_identification, women_in_politics) |>
  filter(party_identification == "Strong Democrat" | party_identification == "Not Strong Democrat" | party_identification == "Independent, Close to Democrat")

republican_data =
  finalized_data |>
  select(year, id, gender, party_identification, women_in_politics) |>
  filter(party_identification == "Strong Republican" | party_identification == "Not Strong Republican" | party_identification == "Independent, Close to Republican")

independent_data =
  finalized_data |>
  select(year, id, gender, party_identification, women_in_politics) |>
  filter(party_identification == "Independent")

other_data =
  finalized_data |>
  select(year, id, gender, party_identification, women_in_politics) |>
  filter(party_identification == "Other")

# Create a specific data set for respondent ages #
age_data = 
  finalized_data |>
  group_by(age) |>
  count() |>
  ungroup() |>
  mutate (pct = percent(n / sum(n), digit = 1)) 

```

# Introduction 
Despite changing norms and perceptions, women are largely still missing from elected office in the United States (U.S.), Canada, and across the world. In 2023, the U.S. elected a historic number of  women to the House of Representatives with 125 (or 28.7%) and 25 to the Senate (or 25%) [@rutgers]. The current Vice President, Kamala Harris, is the first woman to hold this office. Although the number of women elected to the highest levels of government in the U.S. is inching toward gender-parity, progress is slow and hard fought. Women's credibility and perceived power as politicians is impacted by gendered framing by traditional media, priming members of the public to hold biased views about their qualifications [@deGeus]. Framing refers to how information is chosen, interpreted, and presented by the media, influencing the audience’s perception of it as important and shaping their own decision making procedures [@bashevkin]. Previous research has shown that the American public is open to voting for a female president and that gender stereotypes are playing less of a role in the construction of public image and voting behaviours [@StereotypingQs]. However, gender stereotypes continue to be a factor in public perceptions of women in politics. 

To gain a further understanding of the impact of gender stereotypes and the perceived emotional viability of women to serve in politics, I track responses by political views and party identification, using data obtained from the U.S. General Statistical Survey (GSS) from NORC [@gss] at the University of Chicago. These perceptions of women in politics are then analyzed to understand their correlations with demographic factors such as gender and age, as well as tracking the perceptions over time. 

My analysis emphasizes that perceptions of women in politics have generally improved over time since the early 1970s, with a small dip in the late 1990s, 2006, and 2016. People who self-identify as Democrats or as having liberal political views in most cases disagree with the premise that men are emotionally better suited for politics then women, while people who self-identify as Republicans or as having conservative political views in some cases agree with the premise. The number of people with conservative views who agree that men are emotionally better suited for politics is not as high as anticipated, but reinforces the political divide on questions about the role of women in politics. In my Discussion section, I explore these key findings and others of note. 
	
In the remainder of this paper, I commence with the [Data] section which outlines the nature of the data obtained, limitations, and cleaning procedures. In the [Results] section, I present trends found during the analysis process. Then, in the [Discussion] section, I provide further insights and future areas of study. In the [Conclusion], I summarize the main findings. Lastly, the [Appendix] contains a supplemental survey which proposes additional questions to study perceptions of women in politics. 

# Data
The data used in this paper was retrieved from the US General Social Survey (GSS) from NORC at the University of Chicago [@gss]. I obtained both demographic data and data relating to the perceived emotional viability of women in politics, political party affiliation and identification, from 1974 to 2022.

## Source Data

For demographic data, I obtained the following data, as outlined in @tbl-demographic.

| Variable  | New Name | Description      | Example |
|-----------|----------|------------------|---------|
| age       | age      | Respondent's age | 32      |
| sex       | gender   | Respondent's sex | female  |

: Demographic data obtained from GSS {#tbl-demographic}

As noted in @tbl-women_in_pol, I obtained this data relating to the emotional viability of women in serve in politics. The survey question for all three variables asks *"Tell me if you agree or disagree with this statement: Most men are better suited emotionally for politics than are most women"* [@gssCodebook]. Going forward, I will refer to this as the "women in politics question". A further discussion of these variables can be found later on in this section. 

| Variable  | New Name          | Description                                   | Example  |
|-----------|-------------------|-----------------------------------------------|----------|
| fepol     | women_in_politics | Men are better suited for politics than women | agree    |
| fepolv    | women_in_politics | Men are better suited for politics than women | disagree |
| fepolnv   | women_in_politics | Men are better suited for politics than women | disagree |

: Perceptions of Women in Politics data obtained from GSS {#tbl-women_in_pol}

Lastly, as outlined in @tbl-polviews, I retrieved data relating to political views and party identification. The political views survey question asks *"We hear a lot of talk these days about liberals and conservatives. I'm going to show you a seven-point scale on which the political views that people might hold are arranged from extremely liberal--point 1--to extremely conservative--point 7. Where would you place yourself on this scale?"* [@gssCodebook]. The party identification survey question asks *"Generally speaking, do you usually think of yourself as a Republican, Democrat, Independent, or what?"* [@gssCodebook].

| Variable | New Name             | Description                       | Example               |
|----------|----------------------|-----------------------------------|-----------------------|
| polviews | political_views      | Views ranked on a 7-point scale   | Slightly conservative |
| partyid  | party_identification | Republican, Democrat, Independent | Strong Democrat       |

: Political Party and Identification data obtained from GSS {#tbl-polviews}

## Data Limitations
Since 1972, GSS has been conducting in-person interviews to track public opinion. Due to the Covid-19 pandemic, the 2021 and 2022 surveys were conducted solely online and through mixed-methods. This change in methodology resulted in several limitations, which are discussed in further detail below.

### Survey Methodology 
The 2022 GSS bridges the methodology from both the traditional face-to-face data collection processes utilized between 1972 and 2018 and the web-based collection method introduced during the Covid-19 pandemic in 2021, using a mixed mode that includes face-to-face, web, and telephone surveys. Throughout 2022, the GSS conducted an experiment where the sample was divided into two parts. The first part encompassed people doing the face-to-face survey, while the second part included people doing the web and telephone survey. The content of the two parts of the data was not intended to be compared, instead comparing the two modes of collecting data. The sample divided in two for the purpose of comparing methodologies is visible in the `fepolv` and `fepolnv` variables selected for this paper. In 2021 and 2022, the GSS survey updated the variable name from `fepol` to `fepolv` and `fepolnv`, although the phrasing of the question remained the same. Despite these changes in methodology, GSS [@gssCodebook] intends for the 2022 survey to be comparable to the 2018 GSS. 

Previous iterations of GSS selected respondents by quota (1972-1976), Kish grid with age ordering (1975-2018), and last birthday (2021). The 2022 GSS shifts to a Kish grid without age ordering, where adults in the household are selected by their order in the household, not age [@gssCodebook]. These changes to the selection of respondents, especially considering technological literacy among older populations may impact the survey results from 2021 and 2022. 

### Gap in 2020 Data
Data from the year 2020 is not recorded and instead spans 2020 and 2021, which is inconsistent with the typically yearly GSS. There was also a shift in methodology, as previously discussed, to address risks from the Covid-19 pandemic which changed which people were selected and how they completed the GSS questionnaire.

### "No Answer" & "Don't Know" Responses
When moving from an in-person interviewing process to web and mixed-method interviews, GSS shifted their approach to how to code uncertainty, indecision, or a refusal to answer from respondents. Prior to the Covid-19 pandemic, if a respondent opted not to answer a question, interviewers coded it as "No Answer" or "Don't Know". However, with the web-based survey in 2021, there was no interviewer to determine how to code a respondent's non-response. Instead of "No Answer" or "Don't Know", users could skip the question. "Skipped on web" indicates that users read the questions but skipped it. "Skipped on web" responses are separated from “No Answer” which continues to be employed for the face-to-face and phone surveys [@gssCodebook].

### Impact on this study
Although the majority of the data utilized for this study was obtained prior to the Covid-19 pandemic, the 2021 and 2022 responses may be impacted by the changes in methodology and political context during and following the 2020 election. 2020 was a presidential election year in the United States, which saw Joe Biden and Kamala Harris (Democrat) beat Donald Trump and Mike Pence (Republican). Paired with a changing world and policy responses to the Covid-19 pandemic, whether it be mass-vaccination campaigns, stimulus cheques, or relief plans, respondents’ political views and party identification answers may be impacted. As well, Kamala Harris as a woman of colour on the Democrat ticket may have influenced responses to the women in politics survey question. Despite the significant methodological changes and political climate underpinning the 2021 and 2022 GSS, this will have less of an impact on long-term visualizations and trends.

## Data Cleaning
Data was collected, cleaned, and analyzed using the statistical programming software R [@r], using functions from `tidyverse` [@rTidyverse], `ggplot2` [@rGgplot2], `dplyr` [@rDplyr], `janitor` [@rJanitor], `KableExtra` [@rKableExtra], `knitr` [@rknitr], `labelled` [@rLabelled], `haven` [@rHaven], `Formattable` [@rFormattable], and `here` [@rHere]. 

After downloading and selecting the variables of interest from GSS [@gss], I performed data cleaning based on the variable definitions outlined in the GSS codebook [@gssCodebook]. I removed the years in which the questions asking whether men are emotionally better suited for politics than women, political views, and party identification were not asked (the raw data set would display answers for those years as NA). I updated the respondent demographic information data set to reflect the years omitted, given that these demographic questions are asked every year. 

The political views and party identification questions were asked slightly more frequently between 1974 and 1994, including in some years that the women in politics question was not. The responses from 1976, 1980, 1984, and 1987 were removed to ensure consistency between the two data sets. Similar to the women in politics question, the party affiliation and political views were asked bi-annually starting in 1994, except in 2020 due to Covid-19. 

### Respondent Demographic Information
During face-to-face interviews, interviewers traditionally coded the `sex` variable based on their observations. However, it was not possible to make this determination with telephone and web administered surveys, so respondents were asked their sex recorded at birth and their current gender identity. GSS then re-coded these responses into one variable: `sex`. 

During the data cleaning process, I renamed the `sex` column to `gender`, with the value 1 updated to "Male", and the value 2 updated to "Female" based on the GSS codebook [@gssCodebook]. There were no changes in how respondents were asked to provide their age.

### Perceptions of Women in Politics
Between 1974 and 2018, the variable for the women in politics question was `fepol`. However, with the implementation of the online only survey in 2021 and the mixed-methodology survey in 2022, `fepolv` and `fepolnv` were introduced to replace `fepol`. Consequently, employing the `rbind()` function which combines groups of rows together, I merged together the `fepol`, `fepolv`, and `fepolnv` responses into one new variable labelled `women_in_politics`. During the data cleaning process, I update value 1 to "Agree", and value 2 to "Disagree" based on the GSS codebook [@gssCodebook]

There appears to be no pattern or clear methodology behind why the women in politics question was asked some years over others, as sometimes the question was asked two or three years in a row, then not asked in a subsequent year, before repeating the sequence. Starting in 1994, the women in politics question was asked bi-annually, except in 2020 due to Covid-19 impacting the surveying process. 

### Political Preferences
During the data cleaning process, I renamed the `polviews` column to `political_views`, with the value 1 updated to "Extremely liberal", 2 updated to "Liberal", 3 updated to “Slightly liberal,” 4 updated to “Moderate”, 5 updated to “Slightly conservative”, 6 updated to “Conservative”, and 7 to “Extremely conservative” based on the GSS codebook [@gssCodebook].
 
Furthermore, during the data cleaning process I renamed the `partyid` column to `party_identification`. I then changed value 0 to “Strong Democrat", 1 to "Not Strong Democrat", 2 to "Independent, Close to Democrat", 3 to “Independent”, 4 to "Independent, Close to Republican", 5 to "Not Strong Republican", 6 to "Strong Republican", and 7 to “Other", based on the GSS codebook [@gssCodebook]. 

# Results
## Respondent Demographics
Between 1974 and 2022, there were 37,005 responses recorded for GSS surveys which specifically asked the women in politics, political views, and party identification questions. Of these respondents, 20,609 (55.7%) identified as female, with 16,346 (44.2%) identifying as male, and 50 (0.1%) opting not to share their **gender** (see @tbl-gender_summary).

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-gender_summary
#| tbl-cap: Number of respondents by gender 1974-2022

finalized_data |>
  group_by(gender) |>
  count() |>
  ungroup() |>
  mutate (pct = percent(n / sum(n), digit = 1)) |>
  kable(
    col.names = c("Gender", "Number of Responses", "Percentage"),
    booktabs = TRUE
  )
```

The **age** of survey respondents ranged from 18 to 89, with the average respondent being 47 years old (see @tbl-age_summary). 

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-age_summary
#| tbl-cap: Age of Respondents

kable(
  col.names = c("Youngest", "Oldest", "Average Age"),
    booktabs = TRUE,
finalized_data |>
 summarize(
    youngest = min(age, na.rm = TRUE),
    oldest = max(age, na.rm = TRUE),
    average = mean(age, na.rm = TRUE),
 ))
```

Many respondents fell within the 25-43 age range, as visible in @fig-age_summary. @fig-age_summary suggests that the number of teenagers (18 and 19 year olds) who responded to the survey was on the lower end, but quickly rose as respondents entered their 20s and moved toward the age of 25. The number of respondents by age tapered off the older they indicated they were, starting around age 50. 

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: fig-age_summary
#| fig-cap: Number of Respondents by Age

age_data |>
 ggplot(mapping = aes(x = age, y = n)) +
  geom_bar(stat = "identity") +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 25)) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 25)) +
  labs(
    title = "Number of Respondents by Age",
    x = "Age",
    y = "Number of Respondents",
  ) +
  theme_classic() 

```

## Perceptions of Women in Politics 
Despite the biased wording of GSS' women in politics survey question, respondents largely disagreed that men are generally better emotionally suited then women to serve in politics. 74.3% of respondents, spanning 1974-2022, disagree with the question, while 25.7% agree. We can see in @tbl-women_in_pol_summary that women disagree with the question at a higher rate (41.9%), compared to men (32.3%). @tbl-women_in_pol_summary also highlights that women agree with the question at a slightly higher rate than men (13.8% compared to 11.8%). 8 people (0.0%) agreed with the question without sharing their **gender**, while 42 people (0.1%) disagreed. 

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: fig-women_in_pol_year_gender
#| fig-cap: Women in Politics by Year & Gender

finalized_data |>
  ggplot(mapping = aes(x = year, fill = gender)) +
  geom_bar() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=90, size = 6)) + 
  labs(
    x = "Year",
    y = "Number of respondents",
    fill = "Women in Politics"
  ) +
  facet_wrap(vars(women_in_politics)) +
  guides(x = guide_axis(angle = 90)) +
  theme(legend.position = "bottom")

```

@fig-women_in_pol_year_gender illustrates how perceptions of women in politics by **gender** have risen and fallen over time. We can see that both men and women agreed the most frequently with the premise that men are emotionally better suited for politics then women in 1975, 1977, and 1978 and disagreed the most frequently with this premise in 2021 and 2022. The 1985 and 1986 surveys witnessed an increase of both men and women who agree with the women in politics question, then another increase of women who agree with the question in the 1994, 1996, 1998, and 2000 surveys. The number of both men and women who agreed with the women in politics question rose again slightly in 2006 and 2016. Despite slight fluctuations in more recent years, the number of respondents who agree with the premise that men are emotionally better suited for politics has never reached the same levels as when GSS began asking this question in the 1970s. 

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-women_in_pol_summary
#| tbl-cap: Perceptions of Women in Politics

finalized_data |>
  group_by(women_in_politics, gender) |>
  count() |>
  ungroup() |>
  mutate (pct = percent(n / sum(n), digit = 1)) |>
  kable(
    col.names = c("Women in Politics", "Gender", "Number of Responses", "Percentage"),
    booktabs = TRUE
  )

```

The number of respondents by **gender** who disagree with the women in politics question has fluctuated slightly less over time (see @fig-women_in_pol_year_gender). The 1994, 1996, 1998, and 2000 surveys saw an increase in the number of women and men who disagree with the premise that men are emotionally better suited for politics, along with 2006 and 2016. 

Perceptions of women in politics by **age** has also varied over time, as emphasized by @fig-women_in_pol_age. People between the ages of 25 and 45 disagree most frequently with the premise that men are emotionally better suited for politics, with people who identified as being 34 years old disagreeing most strongly. People between the ages of 20 and 25 and 45 and 65 also disagreed with the women in politics question at similar rates. 

Moreover, @fig-women_in_pol_age shows that the number of people who agree with the women in politics question remains relatively the same by age. 

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: fig-women_in_pol_age
#| fig-cap: Women in Politics by Age
finalized_data |>
  ggplot(mapping = aes(x = age, fill = women_in_politics)) +
  geom_bar() +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 25)) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 25)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=90, size = 6)) + 
  labs(
    x = "Age",
    y = "Number of respondents",
    fill = "Women in Politics"
  ) +
  facet_wrap(vars(women_in_politics)) +
  guides(x = guide_axis(angle = 90)) +
  theme(legend.position = "bottom")

```

## Political Views
Responses to the women in politics question varies by **political views**, with people who identify as having more liberal views disagreeing, people with more conservative views agreeing, and people with more moderate views falling in the middle of liberal- and conservative-minded people (see @fig-polviews_all). Liberal views refers to anyone who self-identified in GSS as having `slightly liberal`, `liberal`, or `extremely liberal` views. Conservative views accounts for anyone who self-identified as having `slightly conservative`, `conservative`, or `extremely conservative` views. 
```{r}
#| message: false
#| echo: false
#| label: fig-polviews_all
#| fig-cap: Responses to Women in Politics by All Political Views

#Code for the colour referenced from: https://tellingstorieswithdata.com/05-static_communication.html#bar-charts
finalized_data |>
  ggplot(mapping = aes(x = year, fill = political_views)) +
  geom_bar() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=90, size = 8)) +
  labs(x = "Year", y = "Number of Respondents", fill = "Political Views") +
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(size = 6)) +
  theme(legend.title = element_text(size = 9)) +
  scale_fill_viridis_d(option = "viridis") 

```

As outlined in @tbl-polviews_lib, we can see that 1,965 people (18.6%) who self-identify as holding **liberal** views agree with the premise that men are emotionally better suited for politics, while 8,625 people (81.4%) disagree with the premise. Of the people who agree, 975 (9.2%) identify as female, 988 (9.3%) identify as male, and 2 (0.0%) opted not to identify their **gender**. Of those who disagree with the women in politics question, 4,999 people (47.2%) identify as female, 3,619 identify as male (34.2%), and 7 (0.1%) opted not to identify their **gender**.

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-polviews_lib
#| tbl-cap: Responses to Women in Politics by Political Views- Liberal

liberal_data |>
  group_by(women_in_politics, gender) |>
  count() |>
  ungroup() |>
  mutate (pct = percent(n / sum(n), digit = 1)) |>
  kable(
    col.names = c("Women in Politics", "Gender", "Number of Responses", "Percentage"),
    booktabs = TRUE
  )

```

@tbl-polviews_con highlights that 4,015 people (32.2%) who self-identify as holding **conservative** views agree with the premise of the women in politics question, while 8,472 people (67.8%) who self-identify as holding conservative views disagree with the women in politics question. Of the people who agree, 2,069 (16.6%) identify as female, 1,943 (15.6%) identify as male, and 3 (0.0%) opted not to identify their **gender**. Of those who disagree with the women in politics question, 4,377 people (35.1%) identify as female, 4,079 identify as male (32.7%), and 16 (0.1%) opted not to identify their **gender**.

``` {r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-polviews_con
#| tbl-cap: Responses to Women in Politics by Political Views- Conservative

conservative_data |>
  group_by(women_in_politics, gender) |>
  count() |>
  ungroup() |>
  mutate (pct = percent(n / sum(n), digit = 1)) |>
  kable(
    col.names = c("Women in Politics","Gender", "Number of Responses", "Percentage"),
    booktabs = TRUE
  )

```

Lastly, we can see in @tbl-polviews_moderate that among people who self-identify as holding **moderate** views, 3,523 people (25.3%) agree with the women in politics question while 10,405 people (74.7%) disagree. Of the people who agree, 2,074 (14.9%) identify as female, 1,446 (10.4%) identify as male, and 3 (0.0%) opted not to identify their **gender**. Of those who disagree with the women in politics question, 6,115 people (43.9%) identify as female, 4,271 identify as male (30.7%), and 19 (0.1%) opted not to identify their **gender**.

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-polviews_moderate
#| tbl-cap: Responses to Women in Politics by Political Views- Moderate

moderate_data |>
  group_by(women_in_politics, gender) |>
  count() |>
  ungroup() |>
  mutate (pct = percent(n / sum(n), digit = 1)) |>
  kable(
    col.names = c("Women in Politics", "Gender", "Number of Responses", "Percentage"),
    booktabs = TRUE
  )
```

## Party Identification 
Responses to the question whether men are emotionally better suited for politics then women varies by American political **party identification**. People who self-identify as aligning with the Democrats disagree at a higher percent then people who self-identify as aligning with the Republicans (see @fig-partyid_all). Aligning with the Democrats refers to anyone who self-identified in GSS as `Strong Democrat`, `Not Strong Democrat`, or `Independent, Close to Democrat`. Aligning with the Republicans accounts for anyone who self-identified as `Strong Republican`, `Not Strong Republican`, or `Independent, Close to Republican`. `Independent` refers to people who self-identify with neither the Democrats or Republicans and `Other` accounts for people who align with a third party or candidate. 

```{r}
#| message: false
#| echo: false
#| label: fig-partyid_all
#| fig-cap: Responses to Women in Politics by All Party Identifications

finalized_data |>
  ggplot(mapping = aes(x = year, fill = party_identification)) +
  geom_bar() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=90, size = 8)) +
  labs(x = "Year", y = "Number of Respondents", fill = "Party Identification") +
  theme(legend.position = "bottom") +
  theme(legend.text = element_text(size = 5)) +
  scale_fill_viridis_d(option = "viridis") 

```

@tbl-partyid_dem reveals that 4,146 people (23%) who self-identify as aligning with the **Democrats** agree with the premise that men are emotionally better suited for politics then women, while 13,869 people (77%) who self-identify as being affiliated with the Democrats disagree with the premise. Of the people who agree, 2,286 (12.7%) identify as female, 1,857 (10.3%) identify as male, and 3 (0.0%) opted not to identify their **gender**. Of those who disagree with the women in politics question, 8,278 people (46.0%) identify as female, 5,578 identify as male (31.0%), and 13 (0.1%) opted not to identify their **gender**.

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-partyid_dem
#| tbl-cap: Responses to Women in Politics by Party Identification- Democrat

democrat_data |>
  group_by(women_in_politics, gender) |>
  count() |>
  ungroup() |>
  mutate (pct = percent(n / sum(n), digit = 1)) |>
  kable(
    col.names = c("Women in Politics", "Gender", "Number of Responses", "Percentage"),
    booktabs = TRUE
  )

```

As illustrated by @tbl-partyid_rep, 4,121 people (31.8%) who self-identify as being affiliated with the **Republicans** agree with the women in politics question, while 8,853 people (68.2%) disagree with the question. Of the people who agree, 2,174 (16.8%) identify as female, 1,994 (15.0%) identify as male, and 3 (0.0%) opted not to identify their **gender**. Of those who disagree with the women in politics question, 4,586 people (35.3%) identify as female, 4,256 identify as male (32.8%), and 11 (0.1%) opted not to identify their **gender**.

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-partyid_rep
#| tbl-cap: Responses to Women in Politics by Party Identification- Republican

republican_data |>
  group_by(women_in_politics, gender) |>
  count() |>
  ungroup() |>
  mutate (pct = percent(n / sum(n), digit = 1)) |>
  kable(
    col.names = c("Women in Politics", "Gender", "Number of Responses", "Percentage"),
    booktabs = TRUE
  )

```

@tbl-partyid_ind shows that 1,128 people (20.8%) agree with the women in politics question while 4,294 people (79.2%) who self-identify as being **independent** from a political party disagree. The break down by gender of those who agree is 619 women (11.4%), 507 men (9.4%), and 2 (0.0%) opted not to share their **gender**. Of the people who disagree, 2,423 (44.7%) identify as female, 1,888 (34.2%) identify as male, and 16 (0.3%) opted not to share their **gender**.

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-partyid_ind
#| tbl-cap: Responses to Women in Politics by Party Identification- Independent

independent_data|>
  group_by(women_in_politics, gender) |>
  count() |>
  ungroup() |>
  mutate (pct = percent(n / sum(n), digit = 1)) |>
  kable(
    col.names = c("Women in Politics", "Gender", "Number of Responses", "Percentage"),
    booktabs = TRUE
  )

```

Finally, @tbl-partyid_other demonstrates that 486 people (81.8%) who self-identify as being affiliated with a third party disagree with the women in politics question while 108 people (18.2%) agree with the question. Of the people who agree, 39 (6.6%) identify as female and 69 (11.6%) identify as male. The break down by gender of those who disagree is 204 women (34.3%), 280 men (47.1%), and 2 (0.3%) opted not to share their **gender**. 

```{r}
#| message: false
#| echo: false
#| warning: false
#| label: tbl-partyid_other
#| tbl-cap: Responses to Women in Politics by Party Identification- Other

other_data |>
  group_by(women_in_politics, gender) |>
  count() |>
  ungroup() |>
  mutate (pct = percent(n / sum(n), digit = 1)) |>
  kable(
    col.names = c("Women in Politics", "Gender", "Number of Responses", "Percentage"),
    booktabs = TRUE
  )

```

# Discussion
Overall, the data reveals that despite gendered framing by the traditional media, the American public largely sees women as being emotionally fit to serve as elected officials (74.3% of respondents in @tbl-women_in_pol_summary believe women are emotionally fit). Support for women in politics by gender has risen and fallen over time, potentially responding to election candidates and important news stories of the times (see @fig-women_in_pol_year_gender). With nearly three quarters of respondents disagreeing with the premise that men are more emotionally fit than women for politics, it is important to consider other factors that may be holding us back from achieving gender equality.  

## Perceptions of Women in Politics by Demographic Factors  
As hypothesized, women disagree with the question of whether men are emotionally suited for politics than them at a high rate (see @tbl-women_in_pol_summary). It is noteworthy that women also agree with the question at a slightly higher rate than men (13.8% versus 11.8%). Respondents between the ages of 25 and 45 were identified as being the most supportive of women in politics, which aligns with what is known about the political views and party identifications of people of those ages (see @fig-women_in_pol_age).

## Perceptions of Women in Politics by Political Views 
The number of people with **liberal** views who disagree with the question that men are emotionally better suited for politics then women is comparable to the hypothesized number, while the people with **conservative** views who agree is not as high as anticipated, but still aligns with my hypothesis (see @tbl-polviews_moderate). My hypothesis did not account for people with **moderate** views as the way people with moderate views feel about specific policy issues and survey questions is not consistent nor aligns with a single political party. However, in the case of the women in politics survey question, the people who identified as moderate aligned in the middle of people with liberal and conservative views. 

### Perceptions of Women in Politics - Moderates versus Liberals & Conservatives 
Defining a "moderate" voter in American politics proves challenging and is a weakness of this survey (further discussed in [Limitations]). Scholars view moderates as not simply holding "middle of the road views", rather sharing views with either liberals or conservatives on occasion if the social, economic, or political context aligns with their beliefs [@moderates]. Data from GSS reveals that people with moderate views fall in between people with liberal views and people with conservative views when both agreeing and disagreeing with the women in politics question. We can see in @fig-moderate_liberal that people with moderate views and people with liberal views hold somewhat similar views to the women in politics question (74.7% versus 81.4%). When broken down by gender, 47.2% of liberal-identifying women disagree with the women in politics question (@tbl-polviews_lib), while 43.9% of moderate-identifying women disagree (@tbl-polviews_moderate). 34.2% of liberal-identifying men disagree, while 30.7% of moderate-identifying men disagree. Moreover, 9.2% of liberal-identifying women agree, while 14.9% of moderate-identifying women agree with the premise that men are emotionally better suited for politics. 9.3% of liberal-identifying men agree, while 10.4% of moderate-identifying men agree. In these circumstances, when analyzed by gender, the gap between people with liberal and moderate views closes slightly. 

In contrast, @fig-moderate_con highlights that people with moderate views and people with conservative views also hold somewhat similar beliefs (74.7% versus 67.8%). When broken down by gender, 35.1% of conservative-identifying women disagree with the women in politics question (@tbl-polviews_con), while 43.9% of moderate-identifying women disagree (@tbl-polviews_moderate). 32.7% of conservative-identifying men disagree, while 30.7% of moderate-identifying men disagree. Moreover, 16.6% of conservative-identifying women agree, while 14.9% of moderate-identifying women agree with the premise that men are emotionally better suited for politics. 15.6% of conservative-identifying men agree, while 10.4% of moderate-identifying men agree. When analyzed by gender, the gap between people with conservative and moderate views aligns a little more closely, especially the number of women who agree with the premise that men are emotionally better suited for politics. A larger number of conservative-identifying men agree then moderate-identifying men with the women in politics question. 

While there is often significant disagreement between liberals, conservatives, and sometimes moderates in American politics, it is noteworthy that moderates generally hold "middle of the road views" on this question [@moderates].

```{r}
#| message: false
#| echo: false
#| label: fig-moderate_liberal
#| fig-cap: Comparing Moderate & Liberal Perceptions of Women in Politics

moderate_liberal_data |>
  ggplot(mapping = aes(x = year, fill = political_views)) +
  geom_bar() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=90, size = 5)) + 
  labs(
    x = "Year",
    y = "Women in Politics",
    fill = "Political Views"
  ) +
  facet_wrap(vars(women_in_politics), scales = "free") +
  guides(x = guide_axis(angle = 90)) +
  theme(legend.text = element_text(size = 6)) +
  theme(legend.title = element_text(size = 9)) +
  theme(legend.position = "bottom")

```

```{r}
#| message: false
#| echo: false
#| label: fig-moderate_con
#| fig-cap: Comparing Moderate & Conservative Perceptions of Women in Politics

moderate_conservative_data |>
  ggplot(mapping = aes(x = year, fill = political_views)) +
  geom_bar() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=90, size = 5)) + 
  labs(
    x = "Year",
    y = "Women in Politics",
    fill = "Political Views"
  ) +
  facet_wrap(vars(women_in_politics), scales = "free") +
  guides(x = guide_axis(angle = 90)) +
  theme(legend.text = element_text(size = 6)) +
  theme(legend.title = element_text(size = 9)) +
  theme(legend.position = "bottom") 

```

## Perceptions of Women in Politics by Party Identification 
Despite assumptions and previous research, this GSS data illustrates that perceptions of the emotional viability of women to serve in politics matches political party stances (Democrat and Republican) on the matter, but the results are not as negative and that public is more supportive of women running for elected office then anticipated. The support of women in politics falls along party lines and is more positive than the ratios of women elected to the House of Representatives and the Senate from each party. Currently, there are 92 Democrat women (21%) and 33 Republican women (8%) elected to the House of Representatives (which has 435 members in total) [@rutgers]. There are 15 Democrat women (15%), 9 Republican women (9%), and 1 Independent woman (1%) elected to the Senate (which has 100 members in total) [@rutgers]. 

## Limitations
Asking people about their political views and potential party affiliation remain sensitive topics, despite progress made since the 1970s. As a result, some people may not be comfortable answering these questions which limits the diversity in potential responses collected by GSS. Furthermore, previous research has shown that people may mask their true beliefs when asked about topics such as the viability of women to serve in elected office [@StereotypingQs], which could impact the quality of the data collected and number of "NA" and “don’t know” answers. Ambiguous survey question response options, such as **moderate** (from the political views question) should be clarified to avoid respondents from making an uninformed decision, thus impacting the quality of the data. 

Politics and political views are much more nuanced then simply selecting "agree" or "disagree" on a survey, so without an opportunity for people to explain their views through an open-ended text box, important information which helps contextualize why they do or do not believe women are emotionally viable to serve in elected office is missed. 

Despite the changing variable names to account for newer survey methods, the question asking whether men are emotionally better suited for politics than women hearkens back to the 1970s, despite the strides being made toward gender equality in American politics.  The phrasing of the question asking whether men are better emotionally suited is a leading question which is biased and primes the respondent to think about women in a specific way. Women are framed by traditional media and those on social media as being emotional, high strung, and not possessing “manly” leadership skills, so asking whether men are more emotionally suited over women reinforces traditional media framing and stereotypes [@fallfromgrace] [@bashevkin]. The media never asks men about their emotions or covers their behaviour in the public eye through the lens of being emotional and the same should be expected for the media coverage of women in politics.  GSS should consider asking respondents in future iterations of the survey about leadership qualities, relevant experience, and ability to address crucial policy areas. Questions such as these are addressed in the supplement survey in the [Appendix]. The phrasing of the women in politics survey question does not account for current progress made toward gender equality, nor today's hybrid media environment and the role that social media plays in disseminating political rhetoric and shaping the image of political candidates. It is challenging to know the extent of which these factors impact GSS' data and should be accounted for in upcoming iterations of the survey. 

Finally, GSS' changing sampling methodology over time may have lead to the imbalance of respondents by **gender** (previously seen in @tbl-gender_summary). This imbalance means that people who self-identify as female are represented at a slightly higher rate than people who self-identify as male, meaning readers should be aware when drawing conclusions about gendered perceptions of women in politics. Furthermore, other changes in methodology due to the Covid-19 pandemic may have impacted more recent responses, especially since access to reliable internet, cell phone service, and technological literacy are not a given among all people. 

## Future Research
### Supplemental Survey 
The acceptance of women as political leaders depends on several factors beyond their perceived emotional qualities. Therefore, I prepared a supplemental survey (available in the [Appendix]) to ask respondents about other factors that could impact their views and provide space for further contextualization. 

Previous research reveals that other perceived qualities about political candidates (especially female candidates) such as their leadership skills, highest level of education attained, and their ability to manage "serious" policy issues such as the economy can impact their electoral chances [@fallfromgrace]. Accordingly, the supplemental survey includes questions such as *"Do you think women possess the leadership skills required to serve in elected office?"*, *"Does the gender of a candidate matter more than their policies?"*, and *"What is the most important factor when deciding which candidate to support in an election?"* I hypothesize that perceived leadership skills and the ability to manage important policy issues impact whether respondents decide to vote for female candidates. 

As well, I would like to find out more from respondents, through the question *"Is the current political climate supportive of women seeking elected office?"*, if the political climate is perceived as part of the problem with achieving gender equality. Responses to this question can also inform directions for future studies. 

### Further Studies
There are many socio-economic, cultural, and demographic factors beyond an individual's age and gender that inform voting behaviours and perceptions of what constitutes a good political candidate. Consequently, future research should look at the intersection of race and identity, income, and highest level of education attained with these questions about political views, party identification, and the emotional viability of women to run for office.  

If media framing and political rhetoric are not solely shaping public attitudes towards women in politics, then what is causing the low success rate of electing women to office? Future studies should analyze political party structures, including candidate recruitment, the workplace culture in city halls, state legislatures, the House of Representatives, and the Senate, and the role of social media to help understand and correct course. 

A specific future study should investigate the intersection of political views, party identification, and the emotional viability of women to run for office questions with respondents' voting intentions and voting outcomes during the 2016 presidential election which saw Hillary Clinton (Democrat) face off against Donald Trump (Republican). The GSS variables `vote16`, `pres16`, and `if16who` could provide interesting data and insights [@gssCodebook]. 

# Conclusion
Utilizing data from the U.S. General Social Survey, this paper analyzed perceptions of the emotional viability of women to serve in politics by political views and party identification over time from 1974 to 2022, in correlation with demographic factors such as gender and age. My results reveal that despite gendered media framing, the American public is generally supportive of women seeking elected office. People self-identifying as Democrats and people with liberal and liberal-learning political views are more supportive of women in politics. However, the number of people who self-identify as Republicans and people with conservative and conservative-learning political views were more supportive of women in politics than anticipated. People who self-identified as having Moderate views hold "middle of the road" beliefs in response to the women in politics question, falling approximately between those with self-identified liberal and self-identified conservative views. In future studies, demographic factors such as race, income, and highest level of education attained must be analyzed to understand their intersection with perceptions of women in politics. While media framing and corresponding public perceptions may not be as large of a problem as anticipated, we must work to understand all contributing factors to the low success rate of electing women to the top political offices in the U.S., Canada, and across the world. We cannot let gender equality take another 100 years of hard-fought victories to materialize [@WEF].

# Appendix {.unnumbered}
The supplemental survey is available here: https://forms.gle/EFj72QiGS3fo186z5 

## Survey Preamble 
The General Social Survey collects information and maintains a historical record of respondents' attitudes, experiences, concerns, and practices. In order to strengthen the understanding of perceptions of women in politics, these questions are designed to better understand what socio-economic, cultural, and political factors and conditions contribute to a person's notion of whether women are fit to serve in elected political office, lead efforts to advance specific policy agendas, and increase the number of women elected to all levels of government.

This survey is voluntary and responses are anonymous. If you decide to participate, you can skip questions and withdraw at any time. Individual responses will be recorded in my data set using a user identification number.

Thank you for taking the time to complete this survey. Please contact inessa.deangelis@mail.utoronto.ca if you have questions or require any further information.

## Survey Questions
  (@) If you identify as a visible minority, what is your specific group?

  -Black
  
  -Chinese
  
  -Filipino
  
  -Japanese
  
  -Korean
  
  -South Asian-East Indian
  
  -Southeast Asian
  
  -North African 
  
  -Arab 
  
  -I do not identify as a visible minority
  
  -Other (please specify)
  

(@) What town or city do you live in?

(short answer text)


(@) What state do you live in?

(short answer text)


(@) How often do you engage with politics and political content?

  -Daily
  
  -A few times a week
  
  -A few times a month
  
  -Rarely 
  
  -Never
  
  
  (@) Where do you go to seek out political content?
  
  -Social media
  
  -Traditional media
  
  -Events
  
  -Other
  
  
  (@) Does the gender of a candidate matter more than their political affiliations?
  
  -Yes
  
  -No
  
  -Indifferent
  
  
  (@) Does the gender of a candidate matter more than their political ideology?
  
  -Yes
  
  -No
  
  -Indifferent
  
  
  (@) Does the gender of a candidate matter more than their policies?
  
  -Yes
  
  -No
  
  -Indifferent
  
 
  (@) What is the most important factor when deciding which candidate to support in an election?
  
  -Gender
  
  -Personal characteristics
  
  -Leadership skills
  
  -Ideology
  
  -Political affiliation
  
  -Policies
  
  -Other
  
  
  (@) Do you think women possess the leadership skills required to serve in elected office?
  
  -Yes
  
  -No
  
  
(@) Do you think women can effectively manage political policies?
  
  -Yes
  
  -No
  

(@) Is the current political climate supportive of women seeking elected office?
  
  -Yes
  
  -No
  
  -Somewhat
  
## Submission Message
Thank you for taking the time to complete this survey. Your answers will help expand our understanding of perceptions of women in politics. 

# References
